### 2020-02-17 paper/paper-ancestry-matching.tex
init

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

\renewcommand{\arraystretch}{1.2}

%\usepackage{xr}
%\externaldocument{SCT-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Insights on ancestry using principal component analysis of genetic data}
\author{Florian Priv\'e$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1}}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contact:
\begin{itemize}
\item \url{florian.prive.21@gmail.com}
\end{itemize}

\newpage

\abstract{
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Material and Methods}

\subsection{Ancestry matching \label{ancestry}}

We use the 1000 Genomes (1000G) Project data \cite[]{10002015global} to infer the ancestry of individuals from a study dataset using the following steps: 1) we project individuals from the study dataset onto the PCA space computed using the 1000G data \cite[]{prive2020efficient}; 2) we compute the robust center and covariance parameters of the Mahalanobis distance \cite[]{prive2020efficient} for each of the 26 populations of the 1000G data; 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).

\subsection{Distance between two populations}

To infer a PCA-based genetic distance between two populations, we use the Bhattacharyya distance between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ as defined as
$$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)~,$$
where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}. 
The mean and covariance parameters for each population are computed using the robust center and covariance parameters as in section \ref{ancestry}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

\subsection{Ancestry matching}

We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry.
We did not infer ancestry for 15.9\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.2\%, 11.1\% and 20.3\%, while this represented between 44.5\% (``Chinese'') and 74.3\% (``Bangladeshi'') for other populations. 
In other words, 3 out of 4 of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Mon Oct 21 15:14:34 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1832 & 580 & 31 & 259 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 315 & 13 & 10 & 39 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 155 & 1 & 29 \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 20 &  & 4 \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 34 &  & 4 \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 26 &  &  \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 5 & 469 & 3 & 81 \\ 
   \hline
AMR\_CLM & 1 &  &  & 50 &  &  &  &  &  &  &  &  & 142 \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 20 \\ 
  AMR\_PEL &  &  &  & 2 &  &  &  &  &  &  &  &  & 19 \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 2 &  &  &  &  & 1 \\ 
  EAS\_CHB & 1 &  &  &  &  &  &  & 445 & 10 &  &  &  & 7 \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 380 & 9 &  &  &  & 21 \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 3 & 25 &  &  &  & 122 \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 4 & 16 &  &  &  & 22 \\ 
   \hline
EUR\_CEU & 325725 & 6675 & 350 & 4855 &  &  &  & 1 &  &  &  &  & 1363 \\ 
  EUR\_FIN & 2 &  &  & 118 &  &  &  &  &  &  &  &  & 1 \\ 
  EUR\_GBR & 57497 & 3486 & 79 & 454 &  &  &  &  &  &  & 1 &  & 244 \\ 
  EUR\_IBS & 28 &  & 4 & 452 &  &  &  &  &  &  &  &  & 8 \\ 
  EUR\_TSI & 25 &  & 1 & 177 &  &  &  &  &  &  &  &  & 5 \\ 
   \hline
SAS\_BEB &  &  &  &  & 11 &  & 57 &  &  &  &  &  & 7 \\ 
  SAS\_GIH &  &  &  &  & 334 &  &  &  & 1 &  &  &  & 4 \\ 
  SAS\_ITU & 1 &  &  &  & 314 & 1 &  &  & 264 & 3 &  &  & 82 \\ 
  SAS\_PJL & 1 &  &  &  & 1749 & 767 &  &  & 104 & 2 &  &  & 177 \\ 
  SAS\_STU &  &  &  &  & 38 &  &  &  & 185 &  &  &  & 40 \\ 
   \hline
NA & 47748 & 2594 & 110 & 9705 & 3270 & 980 & 164 & 669 & 1133 & 2140 & 1906 & 73 & 6970 \\ 
   \hline
\end{tabular}
\end{table}

\noindent[OKAY TO CLASSIFY SOME OTHER WHITE AS CENTRAL/SOUTH AMERICANS? MAYBE, SEE NEXT SUBSECTION]

We also  applied our ancestry detection technique to projected individuals from other datasets.
When applied to the European individuals of the POPRES data \cite[]{nelson2008population}, we show that 56.5\% of individuals could not be matched to one of the 26 populations of the 1000G data (Table \ref{tab:ancestry-pred-popres}). 
This is particularly dramatic for all East-Europeans, whose only 11 people could be matched out of 179 (6.1\%).
Moreover, even though 1000G data include several ``Toscani in Italia'' (EUR\_TSI), we were able to match 14.2\% of Italians only.
Nevertheless, all individuals that we could match were identified as of European ancestry. We could also identify accurately sub-regions of Europe; e.g.\ all Spanish and Portugese individuals that we could match (73.1\%) were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
Finally, we also projected and matched people from a case-control cohort for celiac disease composed of four European countries: Italy, the Netherlands, the UK and Finland \cite[]{dubois2010multiple}. 
Among UK individuals, 5397 (79.9\%) were matched to either ``Utah Residents with Northern and Western European Ancestry'' (EUR\_CEU) or ``British in England and Scotland'' (EUR\_GBR), 10 (0.2\%) to other European populations, 1 to ``Mexican Ancestry from Los Angeles USA'' (AMR\_MXL) and 1346 (19.9\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
Among Finns, 1543 (62.4\%) were matched to ``Finnish in Finland'' (EUR\_FIN), 6 (0.3\%) to other European populations, and 922 (37.3\%) could not be matched.
Among Italians, 229 (22\%) were matched to EUR\_TSI, 15 (1.5\%) to EUR\_IBS and 795 (76.5\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).

\noindent[USE FIG S1 TO JUSTIFY 5\% THRESHOLD?]

\subsection{Distance between 1000G populations}

\begin{figure}[!htpb]
	\centerline{\includegraphics[width=0.9\textwidth]{hclust-pop-1000G.png}}
	\caption{Dendrogram of PC-based distances between populations of the 1000G data. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}. \label{fig:ancestry-dist}}
\end{figure}

\noindent[WHAT TO SAY ABOUT THIS? DISTANCE INFORMATIVE? SAS CLOSER TO EUR THAN EAS?]

\noindent[DO THIS IN OTHER DATASETS? COULD SHOW THAT EAST EUROPEAN FAR FROM 1000G POPS IN POPRES. WHAT ABOUT UKBB?]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]

\noindent[DISCUSS PC-BASED GENETIC DISTANCES BETWEEN POPULATIONS]

\noindent[IMPLICATIONS FOR PRS?]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\section*{Software and code availability}

%R package pcadapt is available on CRAN.
%A tutorial on using pcadapt to detect local adaptation is available at \url{https://bcm-uga.github.io/pcadapt/articles/pcadapt.html}.
%The code used in this paper is available at \url{https://github.com/bcm-uga/pcadapt/tree/master/new-paper/code}.

\section*{Acknowledgements}

F.P.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).

\section*{Declaration of Interests}

The authors declare no competing interests.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Supplementary Materials}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage

%\subsection*{Supplementary figures}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\vspace*{1em}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
\label{fig:ancestry-pval}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 00:18:24 2019
\begin{table}[ht]
\centering
\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred-popres}
\begin{tabular}{|l|c|c|c|c|c|}
  \hline
 & EUR\_CEU & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
  \hline
Anglo-Irish Isles & 159 & 57 &  &  & 50 \\ 
  Belgium & 28 & 3 &  &  & 12 \\ 
  Central Europe & 3 &  &  & 1 & 51 \\ 
  Eastern Europe & 2 &  &  &  & 28 \\ 
  France & 14 &  & 13 &  & 62 \\ 
  Germany & 37 &  &  &  & 34 \\ 
  Italy &  &  & 12 & 19 & 188 \\ 
  Netherlands & 10 & 2 &  &  & 5 \\ 
  Scandinavia & 5 &  &  &  & 10 \\ 
  SE Europe &  &  &  & 5 & 89 \\ 
  SW Europe &  &  & 193 &  & 71 \\ 
  Switzerland & 33 & 1 & 6 &  & 182 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 00:28:24 2019
\begin{table}[ht]
\centering
\caption{Ancestry (left) of Celiac individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred-celiac}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline
 & AMR\_MXL & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
  \hline
Finland &  & 4 & 1543 & 1 &  & 1 & 922 \\ 
  Italy &  &  &  &  & 15 & 229 & 795 \\ 
  Netherlands &  & 802 &  & 382 & 1 &  & 463 \\ 
  UK & 1 & 2865 & 1 & 2532 & 4 & 5 & 1346 \\ 
   \hline
\end{tabular}
\end{table}

\end{document}

### 2020-09-21 paper/paper-ancestry-matching.tex
init move code

@@
 
-We use the 1000 Genomes (1000G) Project data \cite[]{10002015global} to infer the ancestry of individuals from a study dataset using the following steps: 1) we project individuals from the study dataset onto the PCA space computed using the 1000G data \cite[]{prive2020efficient}; 2) we compute the robust center and covariance parameters of the Mahalanobis distance \cite[]{prive2020efficient} for each of the 26 populations of the 1000G data; 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
+We use the 1000 Genomes (1000G) Project data \cite[]{10002015global} to infer the ancestry of individuals from a study dataset using the following steps: 1) we project individuals from the study dataset onto the PCA space computed using the 1000G data \cite[]{prive2020efficient}; 2) we compute the robust center and covariance parameters of the Mahalanobis distance \cite[]{prive2020efficient} for each of the 26 populations of the 1000G data; 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we assign each individual the population group with the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
 

### 2020-09-21 code/dist-pop-1000G.R

@@ -1,0 +1,96 @@
+library(bigsnpr)
+library(ggplot2)
+bedfile <- download_1000G("tmp-data")
+plink2 <- download_plink2("tmp-data")
+rel <- snp_plinkKINGQC(
+  plink2.path = plink2,
+  bedfile.in = bedfile,
+  thr.king = 2^-4.5,
+  make.bed = FALSE,
+  ncores = nb_cores()
+)
+str(rel)
+(obj.bed <- bed(bedfile))
+ind.rel <- match(c(rel$IID1, rel$IID2), obj.bed$fam$sample.ID)
+ind.norel <- rows_along(obj.bed)[-ind.rel]
+obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.norel, k = 20,
+                       ncores = nb_cores())
+object.size(obj.bed)
+
+fam2 <- bigreadr::fread2(sub_bed(bedfile, ".fam2"))
+
+U <- obj.svd$u
+pop_PCs <- vctrs::vec_split(U, with(fam2[ind.norel, ],
+                                    paste0(`Super Population`, "_" ,Population)))
+all_maha <- lapply(pop_PCs$val, bigutilsr::covrob_ogk)
+
+bhattacharyya <- function(m1, m2, s1, s2) {
+
+  m_diff <- m2 - m1
+  s_mean <- (s1 + s2) / 2
+
+  drop(crossprod(m_diff, solve(s_mean, m_diff))) / 8 +
+    log(abs(det(s_mean)) / sqrt(abs(det(s1) * det(s2)))) / 2
+}
+
+npop <- length(all_maha)
+dist <- matrix(0, npop, npop)
+for (j in seq_len(npop)) {
+  for (i in seq_len(j - 1)) {
+    dist[i, j] <- dist[j, i] <- bhattacharyya(
+      m1 = all_maha[[i]]$center, m2 = all_maha[[j]]$center,
+      s1 = all_maha[[i]]$cov, s2 = all_maha[[j]]$cov
+    )
+  }
+}
+row.names(dist) <- colnames(dist) <- pop_PCs$key
+dist
+
+hc <- hclust(as.dist(sqrt(dist)))
+plot(hc, xlab = "", ylab = "Distance", col = factor(substr(hc$labels, 1, 3), labels = viridis::viridis(5)))
+
+
+dendrogram <- as.dendrogram(hc)
+library(ggraph)
[…]

### 2020-09-21 code/project-POPRES-to-1000G.R

@@ -1,0 +1,124 @@
+library(bigsnpr)
+library(ggplot2)
+
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = "~/Bureau/POPRES_data/POPRES_allchr",
+  geno = 0.01, mind = 0.01,
+  autosome.only = TRUE
+)
+obj.bed <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
+                         build.new = "hg18", liftOver = "tmp-data/liftOver")
+)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
+    theme_bigstatsr(0.7) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2))
+}))
+
+
+seq_PC <- c(1:18, 20)
+pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
+pop_PCs <- vctrs::vec_split(PC.ref, pop)
+
[…]

### 2020-09-21 code/project-UKBB-to-1000G.R

@@ -1,0 +1,139 @@
+library(bigsnpr)
+library(ggplot2)
+
+obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
+bed.ref <- bed(download_1000G("data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores())
+) # 20 min (including 12 min for projection)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.01 1.02 1.06 1.08 1.36 1.82 2.33 2.36 2.78 2.84 2.99 3.51 4.38
+# [14] 4.67 4.99 5.31 5.74 6.55 6.71 6.75
+
+ind <- sample(nrow(proj1), 20e3)
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[ind, k1], proj1[ind, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[ind, k1], proj2[ind, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+# ggsave("figures/proj1000G-UKBB.png", width = 11, height = 7)
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("data/1000G_phase3_common_norel.fam2")
+
+plot(test$obj.svd.ref)
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
[…]

### 2020-09-21 code/project-celiac-to-1000G.R

@@ -1,0 +1,109 @@
+library(bigsnpr)
+library(ggplot2)
+
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = "~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550",
+  geno = 0.01, mind = 0.01,
+  autosome.only = TRUE
+)
+
+obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
+                         build.new = "hg18", liftOver = "tmp-data/liftOver")
+)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+# ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
+
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
+    theme_bigstatsr(0.7) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2))
+}))
+
+
+seq_PC <- 1:19
[…]

### 2020-09-25 code-save/project-celiac-to-1000G.R
mv unused code

@@ -1,0 +1,128 @@
+library(bigsnpr)
+library(ggplot2)
+
+# snp_plinkQC(
+#   plink.path = download_plink("tmp-data"),
+#   prefix.in = "~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550",
+#   geno = 0.01, mind = 0.01,
+#   autosome.only = TRUE
+# )
+
+obj.bed <- bed("../Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+ind.norel <- readRDS("tmp-data/1008G-ind-norel.rds")
+
+system.time(
+  proj <- bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
+                         k = 25, ncores = nb_cores(),
+                         join_by_pos = FALSE)
+) # 4 min on laptop
+
+PC.ref <- predict(proj$obj.svd.ref)
+proj2 <- proj$OADP_proj
+
+plot_grid(plotlist = lapply(1:9, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+    geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+    theme_bigstatsr(0.5) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2))
+}), nrow = 3)
+
+# ggsave("figures/proj-celiac-to-1000G.png", width = 11, height = 7)
+
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")[ind.norel, ]
+
+source("https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")
+plot_grid2(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
+    theme_bigstatsr(0.7) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2))
+}), title_ratio = 0, legend_ratio = 0.25)
+
+
+seq_PC <- 1:18
+pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
[…]

### 2020-09-25 code/project-POPRES-to-1000G.R
redo POPRES code

@@ -3,122 +3,113 @@ library(ggplot2)
 
 snp_plinkQC(
   plink.path = download_plink("tmp-data"),
-  prefix.in = "~/Bureau/POPRES_data/POPRES_allchr",
+  prefix.in = "../POPRES_data/POPRES_allchr",
   geno = 0.01, mind = 0.01,
   autosome.only = TRUE
 )
-obj.bed <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
+# map <- bigreadr::fread2("../POPRES_data/POPRES_allchr_QC.bim")
+# tab <- bigreadr::fread2("../POPRES_data/POPRES_Snps_QC2.txt", header = FALSE)
+# map$V2 <- tab$V2[match(map$V2, tab$V1)]
+# bigreadr::fwrite2(map, "../POPRES_data/POPRES_allchr_QC.bim")
+
+obj.bed <- bed("../POPRES_data/POPRES_allchr_QC.bed")
 bed.ref <- bed(download_1000G("tmp-data"))
+ind.norel <- readRDS("tmp-data/ind-norel-1000G.rds")
 
-system.time(
-  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
-                         build.new = "hg18", liftOver = "tmp-data/liftOver")
-)
+proj <- bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
+                       k = 25, ncores = nb_cores(),
+                       join_by_pos = FALSE)
+# 1,664,852 variants to be matched.
+# 49,903 ambiguous SNPs have been removed.
+# 275,628 variants have been matched; 49,165 were flipped and 87,081 were reversed.
 
-PC.ref <- predict(test$obj.svd.ref)
-proj1 <- test$simple_proj
-proj2 <- test$OADP_proj
+PC.ref <- predict(proj$obj.svd.ref)
+proj2 <- proj$OADP_proj
 
-
-plot_grid(plotlist = lapply(1:8, function(k) {
+plot_grid(plotlist = lapply(1:9, function(k) {
   k1 <- 2 * k - 1
   k2 <- 2 * k
-  plot_grid(
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
-      theme_bigstatsr(0.5) +
[…]

### 2020-09-26 paper/paper-ancestry-matching.tex
start writing paper again

@@
 
-\title{Insights on ancestry using principal component analysis of genetic data}
-\author{Florian Priv\'e$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1}}$}
+\title{Ancestry inference and grouping\\from principal component analysis of genetic data
+}
+\author{Florian Priv\'e$^{\text{1,}*}$}
 
@@
 
-\section{Introduction}
+$^\dag$ Further defined in supplementary section ``Defintions''.
+
+\section*{Introduction}
+
+Why do we need to group individuals in ancestry groups?
+
+There is no consensus on the method to use to infer population structure, but particularly on grouping individuals.
+
+Why it makes sense to use PCA-based distances -> correction for population structure, and geographical distance.
 
@@
 
-\section{Material and Methods}
+\section*{Measures of genetic dissimilarity between populations}
+
+We first compare four measures of genetic dissimilarity using populations of the 1000 genomes as example \cite[]{10002015global}. 
+The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
+The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
+The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure [TODO]) and provides an appropriate clustering of populations (Figure [TODO]).
+In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures [TODO]).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\section{Material and Methods}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
+\renewcommand{\theequation}{S\arabic{equation}}
+\setcounter{equation}{0}
+\renewcommand{\thesection}{S\arabic{section}}
+\setcounter{section}{1}
+
+\section*{Definitions $\dag$}
+
+\begin{itemize}
+
+	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
+	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.
+
+	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix.
+They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.
+
+	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$, which is defined as
+	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
+	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}.
+	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.
+
+	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points.
+We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
+
+\end{itemize}
 
@@
 We use the 1000 Genomes (1000G) Project data \cite[]{10002015global} to infer the ancestry of individuals from a study dataset using the following steps: 1) we project individuals from the study dataset onto the PCA space computed using the 1000G data \cite[]{prive2020efficient}; 2) we compute the robust center and covariance parameters of the Mahalanobis distance \cite[]{prive2020efficient} for each of the 26 populations of the 1000G data; 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we assign each individual the population group with the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
-
-\subsection{Distance between two populations}
-
-To infer a PCA-based genetic distance between two populations, we use the Bhattacharyya distance between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ as defined as
-$$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)~,$$
-where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}.
-The mean and covariance parameters for each population are computed using the robust center and covariance parameters as in section \ref{ancestry}.
 
@@
 
-\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]
+\noindent[DISCUSS LACK OF SOME POPULATIONS IN 1000G]
 
@@
 
-F.P.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).
+\section*{Funding}
+
+F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).
 

### 2020-09-28 tables/POPRES-matching.tex
code map

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Mon Sep 28 13:09:56 2020
\begin{table}[ht]
\centering
\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
\label{tab:ancestry-pred-popres}
\begin{tabular}{|l|c|c|c|c|c|c|}
  \hline
 & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\
  \hline
Anglo-Irish Isles & 136 &  & 127 &  & 2 & 1 \\
  Belgium & 43 &  &  &  &  &  \\
  Central Europe & 47 &  &  &  & 8 &  \\
  Eastern Europe & 27 &  &  &  & 1 & 2 \\
  France & 49 &  & 3 & 35 & 2 &  \\
  Germany & 67 &  & 3 &  & 1 &  \\
  Italy & 1 &  &  & 11 & 204 & 3 \\
  Netherlands & 13 &  & 4 &  &  &  \\
  Scandinavia & 13 & 1 & 1 &  &  &  \\
  SE Europe & 12 &  &  & 3 & 70 & 9 \\
  SW Europe & 1 &  &  & 261 & 1 & 1 \\
  Switzerland & 179 &  &  & 32 & 11 &  \\
   \hline
\end{tabular}
\end{table}

### 2020-09-28 code/geomap-1000G.R

@@ -1,0 +1,40 @@
+# https://www.internationalgenome.org/data-portal/population
+igsr_pop <- bigreadr::fread2("data/igsr_populations.tsv")
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
+(all_pop <- unique(fam2[3:5]))
+ind <- match(all_pop$Population, igsr_pop$`Population code`)
+df <- data.frame(pop  = all_pop$Population,
+                 desc = all_pop$`Population Description`,
+                 lat  = igsr_pop$`Population latitude`[ind],
+                 lon  = igsr_pop$`Population longitude`[ind])
+
+Amsterdam <- c(52.377956, 4.897070)
+df[df$pop == "CEU", c("lat", "lon")] <- Amsterdam
+Gujarat <- c(22.309425, 72.136230)
+df[df$pop == "GIH", c("lat", "lon")] <- Gujarat
+df[df$pop == "ITU", c("lat", "lon")] <- c(16.5, 79.5)
+df[df$pop == "STU", c("lat", "lon")] <- c(6.927079,	79.861244)
+# Edinburgh <- c(55.953251, -3.188267)
+# London <- c(51.509865, -0.118092)
+# df[df$pop == "GBR", c("lat", "lon")] <- (Edinburgh + London) / 2
+# Mexico <- c(19.432608, -99.133208)
+# df[df$pop == "MXL", c("lat", "lon")] <- Mexico
+# ...
+
+library(leaflet)
+m <- leaflet(df) %>%
+  addTiles() %>%
+  addCircleMarkers(lat = ~lat, lng = ~lon, label = ~paste0(pop, ": ", desc),
+                   radius = 2, color = "red") %>%
+  print()
+
+library(mapview)
+mapshot(m, file = "figures/map-1000G.png")
+
+
+coord <- cbind(df$lon, df$lat)
+geo_dist <- sqrt(apply(coord, 1, geosphere::distGeo, p2 = coord))
+rownames(geo_dist) <- colnames(geo_dist) <- df$pop
+hc <- hclust(as.dist(geo_dist), method = "single")
+heatmap(geo_dist, symm = TRUE, Rowv = as.dendrogram(hc))

### 2020-09-28 code/project-POPRES-to-1000G.R

@@ -50,10 +50,9 @@ plot_grid(plotlist = lapply(7:12, function(k) {
     theme(legend.position = "none") +
     coord_equal()
 }))
+seq_PC <- 1:19
 
 thr_Fst <- 0.002  # Fst between TSI and IBS is 0.0015
-nPC <- 19 # 2, 4, 10, 19
-seq_PC <- seq_len(nPC)
 
 all_centers <- bigutilsr::geometric_median(PC.ref[, seq_PC], by_grp = pop_1000G)
 thr_sq_dist <- thr_Fst * (max(dist(all_centers)^2) / 0.16)
@@ -81,14 +80,14 @@ pop2 <- dplyr::case_when(
   TRUE ~ pop
 )
 
-mean(is.na(choose_pop2)) # 1%
+mean(is.na(choose_pop2)) # 1.2%
 round(100 * tapply(is.na(choose_pop2), pop2, mean), 1)
 # Anglo-Irish Isles           Belgium    Central Europe    Eastern Europe
-#               0.4               0.0               0.0               3.3
+#               0.4               0.0               0.0               6.7
 #            France           Germany             Italy       Netherlands
 #               0.0               0.0               1.4               0.0
 #       Scandinavia         SE Europe         SW Europe       Switzerland
-#               0.0               8.5               0.4               0.0
+#               0.0               9.6               0.4               0.0
 
 library(magrittr)
 table(pop2, choose_pop2, exclude = NULL) %>%
@@ -97,19 +96,19 @@ table(pop2, choose_pop2, exclude = NULL) %>%
   xtable::xtable(caption = "Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \\url{https://www.internationalgenome.org/category/population/}.",
                  label = "tab:ancestry-pred-popres",
                  align = "|l|c|c|c|c|c|c|") %>%
-  print(caption.placement = "top")
+  print(caption.placement = "top", file = "tables/POPRES-matching.tex")
 
 # https://www.internationalgenome.org/category/population/
 #                   EUR_CEU EUR_FIN EUR_GBR EUR_IBS EUR_TSI <NA>
-# Anglo-Irish Isles     132       0     131       0       2    1
+# Anglo-Irish Isles     136       0     127       0       2    1
 # Belgium                43       0       0       0       0    0
-# Central Europe         46       0       0       1       8    0
-# Eastern Europe         28       0       0       0       1    1
[…]

### 2020-10-04 code-save/download-UKBB.R
export code

@@ -1,61 +1,0 @@
-wd <- setwd("data/ukbb_bed")
-for (chr in 1:22) {
-  # bed
-  system(paste0("./ukbgene cal -c", chr))
-  # fam
-  system(paste0("./ukbgene cal -c", chr, " -m"))
-}
-# bim
-download.file("https://biobank.ctsu.ox.ac.uk/crystal/crystal/auxdata/ukb_snp_bim.tar", "ukb_snp_bim.tar")
-untar("ukb_snp_bim.tar")
-
-write(sapply(1:22, function(chr) {
-  c(paste0("ukb_cal_chr", chr, "_v2.bed"),
-    paste0("ukb_snp_chr", chr, "_v2.bim"),
-    paste0("ukb25589_cal_chr", chr, "_v2_s488282.fam"))
-}), tmp <- tempfile(), ncolumns = 3)
-
-library(bigsnpr)
-snp_plinkQC(
-  plink.path = download_plink("."),
-  prefix.in = tmp,
-  file.type = "--merge-list",
-  prefix.out = "ukbb_488282",
-  geno = 0.01,
-  autosome.only = TRUE
-)
-# PLINK v1.90b6.10 64-bit (17 Jun 2019)          www.cog-genomics.org/plink/1.9/
-# (C) 2005-2019 Shaun Purcell, Christopher Chang   GNU General Public License v3
-# Logging to ukbb_488282.log.
-# Options in effect:
-#   --autosome
-#   --geno 0.01
-#   --hwe 1e-50
-#   --maf 0.01
-#   --make-bed
-#   --merge-list /tmp/RtmpHoVtFZ/file32bc56c9faf9b
-#   --mind 0.1
-#   --out ukbb_488282
-#
-# 513440 MB RAM detected; reserving 256720 MB for main workspace.
-# Performing single-pass merge (488377 people, 784256 variants).
-# Merged fileset written to ukbb_488282-merge.bed + ukbb_488282-merge.bim +
-#   ukbb_488282-merge.fam .
-# 784256 variants loaded from .bim file.
-# 488377 people (223474 males, 264808 females, 95 ambiguous) loaded from .fam.
-# Ambiguous sex IDs written to ukbb_488282.nosex .
-# 6 people removed due to missing genotype data (--mind).
-# IDs written to ukbb_488282.irem .
-# Using 1 thread (no multithreaded calculations invoked).
-# Before main variant filters, 488371 founders and 0 nonfounders present.
-# Calculating allele frequencies... done.
[…]

### 2020-10-04 code-save/project-celiac-to-1000G.R

@@ -1,128 +1,0 @@
-library(bigsnpr)
-library(ggplot2)
-
-# snp_plinkQC(
-#   plink.path = download_plink("tmp-data"),
-#   prefix.in = "~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550",
-#   geno = 0.01, mind = 0.01,
-#   autosome.only = TRUE
-# )
-
-obj.bed <- bed("../Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
-bed.ref <- bed(download_1000G("tmp-data"))
-ind.norel <- readRDS("tmp-data/1008G-ind-norel.rds")
-
-system.time(
-  proj <- bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
-                         k = 25, ncores = nb_cores(),
-                         join_by_pos = FALSE)
-) # 4 min on laptop
-
-PC.ref <- predict(proj$obj.svd.ref)
-proj2 <- proj$OADP_proj
-
-plot_grid(plotlist = lapply(1:9, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-    geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
-    theme_bigstatsr(0.5) +
-    labs(x = paste0("PC", k1), y = paste0("PC", k2))
-}), nrow = 3)
-
-# ggsave("figures/proj-celiac-to-1000G.png", width = 11, height = 7)
-
-
-#### Ancestry estimation ####
-
-fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")[ind.norel, ]
-
-source("https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")
-plot_grid2(plotlist = lapply(5:10, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
-    theme_bigstatsr(0.7) +
-    labs(x = paste0("PC", k1), y = paste0("PC", k2))
-}), title_ratio = 0, legend_ratio = 0.25)
-
-
-seq_PC <- 1:18
-pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
[…]

### 2020-10-04 code/ancestry-grouping.R

@@ -1,0 +1,226 @@
+library(bigsnpr)
+library(ggplot2)
+NCORES <- 15
+
+#### Groupings of homogeneous ancestry  ####
+
+## 1) Matching to 1000G populations
+
+# Project UKBB individuals onto PCA space from 1000G
+
+obj.bed <- bed("data/ukbb.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+ind.norel <- readRDS("tmp-data/ind-norel-1000G.rds")
+
+proj <- runonce::save_run(
+  bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
+                 k = 25, ncores = NCORES),
+  file = "data/proj-UKBB-to-1000G.rds"
+) # < 15 min
+# 1,664,852 variants to be matched.
+# 22,794 ambiguous SNPs have been removed.
+# 377,914 variants have been matched; 0 were flipped and 65,346 were reversed.
+
+PC.ref <- predict(proj$obj.svd.ref)
+proj2 <- proj$OADP_proj
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")[ind.norel, ]
+pop_1000G <- paste0(fam2$`Super Population`, "_", fam2$Population)
+
+seq_PC <- 1:16
+thr_Fst <- 0.002
+
+all_centers <- bigutilsr::geometric_median(PC.ref[, seq_PC], by_grp = pop_1000G)
+thr_sq_dist <- thr_Fst * (max(dist(all_centers)^2) / 0.16)
+
+all_sq_dist <- apply(all_centers, 1, function(center) {
+  bigutilsr:::rowSumsSq(sweep(proj2[, seq_PC], 2, center, '-'))
+})
+
+## British ancestry (and around)
+qplot(all_sq_dist[, "EUR_GBR"]) +
+  theme_bigstatsr(0.9) +
+  scale_x_log10() +
+  labs(x = "Squared distance to GBR cluster") +
+  geom_vline(xintercept = thr_sq_dist, color = "red")
+# ggsave("figures/hist-dist-GBR.pdf", width = 7, height = 5)
+
+mean(all_sq_dist[, "EUR_GBR"] < thr_sq_dist) # 89.73%
+ind_GBR <- which(all_sq_dist[, "EUR_GBR"] < thr_sq_dist)
+
+df0 <- readRDS("data/info_UKBB.rds")
[…]

### 2020-10-04 code/dist-pop-1000G.R

@@ -1,96 +1,171 @@
 library(bigsnpr)
-library(ggplot2)
+NCORES <- 15
+
+#### Compute PCA on 1000G ####
+
 bedfile <- download_1000G("tmp-data")
-plink2 <- download_plink2("tmp-data")
-rel <- snp_plinkKINGQC(
-  plink2.path = plink2,
-  bedfile.in = bedfile,
-  thr.king = 2^-4.5,
-  make.bed = FALSE,
-  ncores = nb_cores()
-)
-str(rel)
 (obj.bed <- bed(bedfile))
-ind.rel <- match(c(rel$IID1, rel$IID2), obj.bed$fam$sample.ID)
-ind.norel <- rows_along(obj.bed)[-ind.rel]
-obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.norel, k = 20,
-                       ncores = nb_cores())
-object.size(obj.bed)
+plink2 <- download_plink2("tmp-data")
 
-fam2 <- bigreadr::fread2(sub_bed(bedfile, ".fam2"))
+ind.norel <- runonce::save_run({
+  rel <- snp_plinkKINGQC(
+    plink2.path = plink2,
+    bedfile.in = bedfile,
+    thr.king = 2^-4.5,
+    make.bed = FALSE,
+    ncores = NCORES,
+    extra.options = "--memory 5000"
+  )
+  str(rel)
+  which(!obj.bed$fam$sample.ID %in% rel$IID2)
+}, file = "tmp-data/ind-norel-1000G.rds")
 
-U <- obj.svd$u
-pop_PCs <- vctrs::vec_split(U, with(fam2[ind.norel, ],
-                                    paste0(`Super Population`, "_" ,Population)))
-all_maha <- lapply(pop_PCs$val, bigutilsr::covrob_ogk)
+obj.svd <- runonce::save_run(
+  bed_autoSVD(obj.bed, ind.row = ind.norel, k = 25, ncores = NCORES),
+  file = "tmp-data/SVD-1000G.rds"
+)
+plot(obj.svd, type = "scores", scores = 14:25, coeff = 0.6)
+PC <- predict(obj.svd)[, 1:19]
+
+
+#### Compute multiple distances between 1000G populations ####
+
+fam2 <- bigreadr::fread2(sub_bed(bedfile, ".fam2"))[ind.norel, ]
+pop_1000G <- fam2$Population
+split_pop <- split(rows_along(fam2), pop_1000G)
+npop <- length(split_pop)
+
+## Fst
+
[…]

### 2020-10-04 code/prepare-UKBB.R

@@ -1,0 +1,52 @@
+##### Prepare genotypes ####
+
+# file.symlink("~/NCRR-PRS/faststorage/UKBB/", ".")
+
+write(sapply(1:22, function(chr) {
+  paste0("UKBB/bed/",
+         c(paste0("ukb_cal_chr", chr, "_v2.bed"),
+           paste0("ukb_snp_chr", chr, "_v2.bim"),
+           paste0("ukb58024_cal_chr", chr, "_v2_s488264.fam")))
+}), tmp <- tempfile(), ncolumns = 3)
+
+library(bigsnpr)
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = tmp,
+  file.type = "--merge-list",
+  prefix.out = "data/ukbb",
+  geno = 0.01,
+  autosome.only = TRUE,
+  extra.options = "--memory 100000"
+)
+
+#### Prepare other data (self-reported ancestry and PCs) ####
+
+library(bigreadr)
+library(dplyr)
+
+## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
+code_ancestry <- fread2("UKBB/coding1001.tsv")
+unknown <- c("Prefer not to answer", "Do not know", "Mixed",
+             "Other ethnic group", "Any other mixed background")
+
+df0 <- fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "21000-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "pop", paste0("PC", 1:16))
+) %>%
+  mutate(
+    pop  = factor(pop, levels = code_ancestry$coding,
+                  labels = code_ancestry$meaning) %>%
+      forcats::fct_recode("Other White" = "Any other white background",
+                          "Other Asian" = "Any other Asian background",
+                          "Other Black" = "Any other Black background") %>%
+      forcats::fct_other(drop = unknown, other_level = NA) %>%
+      droplevels(pop, exclude = NA) %>%
+      forcats::fct_relevel(c(
+        "British", "Irish", "White", "Other White",
+        "Indian", "Pakistani", "Bangladeshi", "Chinese", "Other Asian",
+        "Caribbean", "African", "Other Black"))
+  )
+str(df0)
+saveRDS(df0, "data/info_UKBB.rds")

### 2020-10-04 code/project-UKBB-to-1000G.R

@@ -1,139 +1,229 @@
 library(bigsnpr)
 library(ggplot2)
+NCORES <- 15
 
-obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
-bed.ref <- bed(download_1000G("data"))
+#### Project UKBB individuals onto PCA space from 1000G ####
 
-system.time(
-  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores())
-) # 20 min (including 12 min for projection)
+obj.bed <- bed("data/ukbb.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+ind.norel <- readRDS("tmp-data/ind-norel-1000G.rds")
 
-PC.ref <- predict(test$obj.svd.ref)
-proj1 <- test$simple_proj
-proj2 <- test$OADP_proj
+proj <- runonce::save_run(
+  bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
+                 k = 25, ncores = NCORES),
+  file = "data/proj-UKBB-to-1000G.rds"
+) # < 15 min
+# 1,664,852 variants to be matched.
+# 22,794 ambiguous SNPs have been removed.
+# 377,914 variants have been matched; 0 were flipped and 65,346 were reversed.
 
-# shrinkage coefficients
-shrinkage <- unname(sapply(1:20, function(k) {
-  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
-}))
-round(shrinkage, 2)
-#  [1] 1.01 1.02 1.06 1.08 1.36 1.82 2.33 2.36 2.78 2.84 2.99 3.51 4.38
-# [14] 4.67 4.99 5.31 5.74 6.55 6.71 6.75
+PC.ref <- predict(proj$obj.svd.ref)
+proj2 <- proj$OADP_proj
 
-ind <- sample(nrow(proj1), 20e3)
+ind <- sample(nrow(proj2), 20e3)
+# ind <- rows_along(proj2)
 
-plot_grid(plotlist = lapply(1:8, function(k) {
+plot_grid(plotlist = lapply(1:9, function(k) {
   k1 <- 2 * k - 1
   k2 <- 2 * k
-  plot_grid(
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj1[ind, k1], proj1[ind, k2]), color = "red") +
[…]

### 2020-10-04 paper/paper-ancestry-matching.tex
continue writing paper

@@
 \usepackage{makecell}
+\usepackage{placeins}
 
@@
 
-\section*{Introduction}
+\subsection*{Introduction}
 
@@
 
-Why it makes sense to use PCA-based distances -> correction for population structure, and geographical distance.
+Why it makes sense to use PCA-based distances -> correction for population structure, and geographical distance. + Fast
+
+Global ancestry only
 
@@
 
-\section*{Measures of genetic dissimilarity between populations}
+\subsection*{Measures of genetic dissimilarity between populations}
 
@@
 The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
+We report $F_{ST}$ between the 26 1000G populations in tables S1-S5.
 The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
-The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure [TODO]) and provides an appropriate clustering of populations (Figure [TODO]).
-In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures [TODO]).
+The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
+In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1} and \ref{fig:heatmap3}).
+For example, using , African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
+Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.
 
@@
 
-%\section{Material and Methods}
+\subsection*{PCA-based ancestry inference}
 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-\renewcommand{\theequation}{S\arabic{equation}}
-\setcounter{equation}{0}
-\renewcommand{\thesection}{S\arabic{section}}
-\setcounter{section}{1}
+We project the dataset of interest (either UKBB or POPRES) onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
+We recall that this uses a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}. 
+Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the minimum distance to their respective center (geometric median$^\dag$).
+Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.
+0.002.
+This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
+When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown.
 
-\section*{Definitions $\dag$}
-
-\begin{itemize}
-
-	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
-	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.
-
-	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix.
-They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.
-
-	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$, which is defined as
-	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
-	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}.
-	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.
-
-	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points.
-We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
-
-\end{itemize}
-
-\subsection{Ancestry matching \label{ancestry}}
-
-We use the 1000 Genomes (1000G) Project data \cite[]{10002015global} to infer the ancestry of individuals from a study dataset using the following steps: 1) we project individuals from the study dataset onto the PCA space computed using the 1000G data \cite[]{prive2020efficient}; 2) we compute the robust center and covariance parameters of the Mahalanobis distance \cite[]{prive2020efficient} for each of the 26 populations of the 1000G data; 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we assign each individual the population group with the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-\subsection{Ancestry matching}
-
+[TODO: CHANGE NUMBERS]
 We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry.
@@
 Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
-
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Mon Oct 21 15:14:34 2019
-\begin{table}[ht]
-\centering
-\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred}
-\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-  \hline
- & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
-  \hline
-AFR\_ACB &  &  &  &  &  &  &  &  &  & 1832 & 580 & 31 & 259 \\ 
-  AFR\_ASW &  &  &  &  &  &  &  &  &  & 315 & 13 & 10 & 39 \\ 
-  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 155 & 1 & 29 \\ 
-  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 20 &  & 4 \\ 
-  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 34 &  & 4 \\ 
-  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 26 &  &  \\ 
-  AFR\_YRI &  &  &  &  &  &  &  &  &  & 5 & 469 & 3 & 81 \\ 
-   \hline
-AMR\_CLM & 1 &  &  & 50 &  &  &  &  &  &  &  &  & 142 \\ 
-  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 20 \\ 
-  AMR\_PEL &  &  &  & 2 &  &  &  &  &  &  &  &  & 19 \\ 
-   \hline
-EAS\_CDX &  &  &  &  &  &  &  & 2 &  &  &  &  & 1 \\ 
-  EAS\_CHB & 1 &  &  &  &  &  &  & 445 & 10 &  &  &  & 7 \\ 
-  EAS\_CHS &  &  &  &  &  &  &  & 380 & 9 &  &  &  & 21 \\ 
-  EAS\_JPT &  &  &  &  &  &  &  & 3 & 25 &  &  &  & 122 \\ 
-  EAS\_KHV &  &  &  &  &  &  &  & 4 & 16 &  &  &  & 22 \\ 
-   \hline
-EUR\_CEU & 325725 & 6675 & 350 & 4855 &  &  &  & 1 &  &  &  &  & 1363 \\ 
-  EUR\_FIN & 2 &  &  & 118 &  &  &  &  &  &  &  &  & 1 \\ 
-  EUR\_GBR & 57497 & 3486 & 79 & 454 &  &  &  &  &  &  & 1 &  & 244 \\ 
-  EUR\_IBS & 28 &  & 4 & 452 &  &  &  &  &  &  &  &  & 8 \\ 
-  EUR\_TSI & 25 &  & 1 & 177 &  &  &  &  &  &  &  &  & 5 \\ 
-   \hline
-SAS\_BEB &  &  &  &  & 11 &  & 57 &  &  &  &  &  & 7 \\ 
-  SAS\_GIH &  &  &  &  & 334 &  &  &  & 1 &  &  &  & 4 \\ 
-  SAS\_ITU & 1 &  &  &  & 314 & 1 &  &  & 264 & 3 &  &  & 82 \\ 
-  SAS\_PJL & 1 &  &  &  & 1749 & 767 &  &  & 104 & 2 &  &  & 177 \\ 
-  SAS\_STU &  &  &  &  & 38 &  &  &  & 185 &  &  &  & 40 \\ 
-   \hline
-NA & 47748 & 2594 & 110 & 9705 & 3270 & 980 & 164 & 669 & 1133 & 2140 & 1906 & 73 & 6970 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\noindent[OKAY TO CLASSIFY SOME OTHER WHITE AS CENTRAL/SOUTH AMERICANS? MAYBE, SEE NEXT SUBSECTION]
 
@@
 
-\noindent[USE FIG S1 TO JUSTIFY 5\% THRESHOLD?]
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Sun Sep 27 11:26:46 2020
+\begin{table}[ht]
+	\centering
+	\caption{Self-reported ancestry (left) of UKBB individuals and their matching to 1000G continental populations (top) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+	\label{tab:infer-UKBB-superpop}
+	\begin{tabular}{|l|c|c|c|c|c|c|}
+		\hline
+		& AFR & AMR & EAS & EUR & SAS & Not matched \\ 
+		\hline
+		British & 2 &  & 1 & 421457 & 6 & 9548 \\ 
+		Irish &  &  &  & 12328 &  & 425 \\ 
+		White & 1 & 1 & 1 & 499 &  & 43 \\ 
+		Other White &  & 40 &  & 11334 & 1 & 4440 \\ 
+		\hline
+		Indian &  &  &  & 5 & 4922 & 789 \\ 
+		Pakistani &  &  &  &  & 1421 & 327 \\ 
+		Bangladeshi &  &  &  &  & 217 & 4 \\ 
+		Chinese &  &  & 1453 & 1 &  & 50 \\ 
+		Other Asian & 1 &  & 279 &  & 939 & 528 \\ 
+		\hline
+		Caribbean & 3848 &  &  &  & 25 & 424 \\ 
+		African & 2633 &  &  & 1 &  & 570 \\ 
+		Other Black & 74 &  &  &  & 2 & 42 \\ 
+		\hline
+		Asian or Asian British &  &  & 2 &  & 20 & 20 \\ 
+		Black or Black British & 20 &  &  & 2 &  & 4 \\ 
+		White and Black Caribbean & 24 & 1 &  & 8 & 1 & 563 \\ 
+		White and Black African & 5 &  &  & 6 &  & 391 \\ 
+		White and Asian &  & 1 & 2 & 27 & 26 & 746 \\ 
+		\hline
+		Unknown  & 835 & 173 & 576 & 2296 & 633 & 3307 \\ 
+		\hline
+	\end{tabular}
+\end{table}
 
-\subsection{Distance between 1000G populations}
-
-\begin{figure}[!htpb]
-	\centerline{\includegraphics[width=0.9\textwidth]{hclust-pop-1000G.png}}
-	\caption{Dendrogram of PC-based distances between populations of the 1000G data.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.
-\label{fig:ancestry-dist}}
-\end{figure}
-
-\noindent[WHAT TO SAY ABOUT THIS? DISTANCE INFORMATIVE? SAS CLOSER TO EUR THAN EAS?]
-
-\noindent[DO THIS IN OTHER DATASETS? COULD SHOW THAT EAST EUROPEAN FAR FROM 1000G POPS IN POPRES. WHAT ABOUT UKBB?]
+We also test the approach proposed in \cite{zhang2020fast} which consists of finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
+When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
 
@@
 
-\section{Discussion}
+\subsection*{PCA-based ancestry grouping}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
+	\caption{ \label{fig:grouping}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{Discussion}
+
+Global measures vs relative measures.
+
+Contrast with Mahalanobis distance proposed before.
 
@@
 
-\newpage
 \section*{Supplementary Materials}
@@
 \setcounter{table}{0}
+\renewcommand{\theequation}{S\arabic{equation}}
+\setcounter{equation}{0}
+\renewcommand{\thesection}{S\arabic{section}}
+\setcounter{section}{1}
+
+\subsection*{Definitions $\dag$}
+
+\begin{itemize}
+
+	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
+	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.
+
+	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix.
+They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.
+
+	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$, which is defined as
+	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
+	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}.
+	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.
+
+	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points.
+We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
+
+\end{itemize}
 
@@
 
-%\clearpage
+\clearpage
 
-%\subsection*{Supplementary figures}
+\subsection*{Location of 1000G populations}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\vspace*{1em}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
-\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
-\label{fig:ancestry-pval}}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{map-1000G}}
+	\caption{ \label{fig:map}}
 \end{figure}
@@
 
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 00:18:24 2019
-\begin{table}[ht]
-\centering
-\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred-popres}
-\begin{tabular}{|l|c|c|c|c|c|}
-  \hline
- & EUR\_CEU & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
-  \hline
-Anglo-Irish Isles & 159 & 57 &  &  & 50 \\ 
-  Belgium & 28 & 3 &  &  & 12 \\ 
-  Central Europe & 3 &  &  & 1 & 51 \\ 
-  Eastern Europe & 2 &  &  &  & 28 \\ 
-  France & 14 &  & 13 &  & 62 \\ 
-  Germany & 37 &  &  &  & 34 \\ 
-  Italy &  &  & 12 & 19 & 188 \\ 
-  Netherlands & 10 & 2 &  &  & 5 \\ 
-  Scandinavia & 5 &  &  &  & 10 \\ 
-  SE Europe &  &  &  & 5 & 89 \\ 
-  SW Europe &  &  & 193 &  & 71 \\ 
-  Switzerland & 33 & 1 & 6 &  & 182 \\ 
-   \hline
-\end{tabular}
-\end{table}
+\clearpage
 
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 00:28:24 2019
-\begin{table}[ht]
-\centering
-\caption{Ancestry (left) of Celiac individuals and their matching to 1000G populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred-celiac}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-  \hline
- & AMR\_MXL & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
-  \hline
-Finland &  & 4 & 1543 & 1 &  & 1 & 922 \\ 
-  Italy &  &  &  &  & 15 & 229 & 795 \\ 
-  Netherlands &  & 802 &  & 382 & 1 &  & 463 \\ 
-  UK & 1 & 2865 & 1 & 2532 & 4 & 5 & 1346 \\ 
-   \hline
-\end{tabular}
-\end{table}
+\subsection*{Measures of genetic dissimilarity between populations}
+
+\input{../tables/fst-values-AFR}
+\input{../tables/fst-values-AMR}
+\input{../tables/fst-values-EAS}
+\input{../tables/fst-values-EUR}
+\input{../tables/fst-values-SAS}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-dist-to-Fst}}
+	\caption{Comparing Fst to different distances in the PCA space between pairs of the 26 1000G populations. \label{fig:compare-dist}}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=0.8\textwidth]{heatmap-Fst-1000G}}
+\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in table S1-S5. \label{fig:heatmap0}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-bhattacharyya-1000G}}
+	\caption{Heatmap with clustering based on the Bhattacharyya distances between pairs of the 26 1000G populations. \label{fig:heatmap1}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-centers-1000G}}
+	\caption{Heatmap with clustering based on the distances between centers of pairs of the 26 1000G populations. \label{fig:heatmap2}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-closest-1000G}}
+	\caption{Heatmap with clustering based on the shortest distances between pairs of the 26 1000G populations. \label{fig:heatmap3}}
+\end{figure}
+
+\FloatBarrier
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{PCA-based ancestry inference}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{proj-UKBB-to-1000G}}
+	\caption{ \label{fig:proj-UKBB}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-mismatch}}
+	\caption{ \label{fig:mismatch}}
+\end{figure}
+
+\input{../tables/matched-pop-KNN}
+\input{../tables/infer-UKBB-pop}
+\input{../tables/POPRES-matching}
+
+\FloatBarrier
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{PCA-based ancestry grouping}
+
+\input{../tables/ancestry-groups}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
+	\caption{ \label{fig:hist1}}
+\end{figure}
+\begin{figure}[h]
+
+\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
+\caption{ \label{fig:hist2}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
+	\caption{ \label{fig:hist3}}
+\end{figure}
+
+\FloatBarrier
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-10-05 code/ancestry-grouping.R
match before

@@ -48,9 +48,8 @@ qplot(all_sq_dist[, "EUR_GBR"]) +
 mean(all_sq_dist[, "EUR_GBR"] < thr_sq_dist) # 89.73%
 ind_GBR <- which(all_sq_dist[, "EUR_GBR"] < thr_sq_dist)
 
-df0 <- readRDS("data/info_UKBB.rds")
-df0 <- df0[match(obj.bed$fam$sample.ID, df0$eid), ]
-pop_UKBB <- df0$pop
+df <- readRDS("data/info_UKBB.rds")
+pop_UKBB <- df$pop
 table(as.character(pop_UKBB[ind_GBR]), exclude = NULL)
 
 ## African ancestry
@@ -86,16 +85,16 @@ ind_pop <- list(ind_GBR, ind_AFR, ind_EAS, ind_SAS, ind_AMR)
 length(unlist(ind_pop)) # 456410
 length(unique(unlist(ind_pop))) # 456410
 matched_pop <- rep(c("GBR", "AFR", "EAS", "SAS", "AMR"), lengths(ind_pop))
-df0_pop <- df0[unlist(ind_pop), ]
+df_pop <- df[unlist(ind_pop), ]
 
 plot_grid(
-  qplot(PC1, PC2, data = df0_pop, color = matched_pop) +
+  qplot(PC1, PC2, data = df_pop, color = matched_pop) +
     theme_bigstatsr(0.8) +
     labs(color = "Ancestry") +
     theme(legend.position = c(0.75, 0.4)) +
     guides(colour = guide_legend(override.aes = list(size = 2))),
 
-  qplot(PC3, PC4, data = df0_pop, color = matched_pop) +
+  qplot(PC3, PC4, data = df_pop, color = matched_pop) +
     theme_bigstatsr(0.8) +
     theme(legend.position = "none"),
 
@@ -108,7 +107,7 @@ plot_grid(
 ## 2) Using self-reported ancestry (without projection)
 
 library(dplyr)
-PC_UKBB <- as.matrix(select(df0, PC1:PC16))
+PC_UKBB <- as.matrix(select(df, PC1:PC16))
 mixed <- c("Asian or Asian British", "Black or Black British",
            "White and Black Caribbean", "White and Black African",
            "White and Asian")

### 2020-10-05 code/prepare-UKBB.R

@@ -49,4 +49,9 @@ df0 <- fread2(
         "Caribbean", "African", "Other Black"))
   )
 str(df0)
-saveRDS(df0, "data/info_UKBB.rds")
+
+bed_eid <- fread2("data/ukbb.fam")[[2]]
+df <- df0[match(bed_eid, df0$eid), ]
+mean(is.na(df$pop))  # 1.6%
+
+saveRDS(df, "data/info_UKBB.rds")

### 2020-10-05 code/project-UKBB-to-1000G.R

@@ -71,8 +71,8 @@ choose_pop2 <- ifelse(min_sq_dist > thr_sq_dist, NA,
 mean(is.na(choose_pop2)) # 4.55%
 
 
-df0 <- readRDS("data/info_UKBB.rds")
-pop_UKBB <- df0$pop[match(obj.bed$fam$sample.ID, df0$eid)]
+df <- readRDS("data/info_UKBB.rds")
+pop_UKBB <- df$pop
 mixed <- c("Asian or Asian British", "Black or Black British",
            "White and Black Caribbean", "White and Black African",
            "White and Asian")
@@ -144,7 +144,7 @@ ind_mistmatch <- c(
   which(pop_UKBB %in% c("Caribbean", "African", "Other Black") &
           substr(choose_pop2, 1, 3) != "AFR")  # 28, mainly Caribbean as SAS
 ) # 47 values
-PC_UKBB <- df0[match(obj.bed$fam$sample.ID, df0$eid), -(1:2)]
+PC_UKBB <- df[, -(1:2)]
 
 ind <- sample(which(!is.na(pop_UKBB2)), 50e3)
 

### 2020-10-05 paper/paper-ancestry-matching.tex
write more results

@@
 
-\newpage
+\vspace*{5em}
 
@@
 
-[TODO: CHANGE NUMBERS]
-We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry.
-We did not infer ancestry for 15.9\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
-More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.2\%, 11.1\% and 20.3\%, while this represented between 44.5\% (``Chinese'') and 74.3\% (``Bangladeshi'') for other populations.
-In other words, 3 out of 4 of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
-Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
-Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
+We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
+These individuals seem to originate from all over the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
+Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
+Based on the threshold defined before, we could not infer ancestry for 4.6\% of the individuals.
+More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}).
+This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries.
+Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g. 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian`.
+Only 47 individuals were wrongly classified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
+However, when comparing the location of these individuals on the PCA space computed in the UK Biobank \cite[]{bycroft2017genome} to the rest of individuals, it seemed more probable that the self-reported ancestry is wrong and not our ancestry estimate (Figure \ref{fig:mismatch}).
 
-We also  applied our ancestry detection technique to projected individuals from other datasets.
-When applied to the European individuals of the POPRES data \cite[]{nelson2008population}, we show that 56.5\% of individuals could not be matched to one of the 26 populations of the 1000G data (Table \ref{tab:ancestry-pred-popres}).
-This is particularly dramatic for all East-Europeans, whose only 11 people could be matched out of 179 (6.1\%).
-Moreover, even though 1000G data include several ``Toscani in Italia'' (EUR\_TSI), we were able to match 14.2\% of Italians only.
-Nevertheless, all individuals that we could match were identified as of European ancestry.
-We could also identify accurately sub-regions of Europe; e.g.\ all Spanish and Portugese individuals that we could match (73.1\%) were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
-Finally, we also projected and matched people from a case-control cohort for celiac disease composed of four European countries: Italy, the Netherlands, the UK and Finland \cite[]{dubois2010multiple}.
-Among UK individuals, 5397 (79.9\%) were matched to either ``Utah Residents with Northern and Western European Ancestry'' (EUR\_CEU) or ``British in England and Scotland'' (EUR\_GBR), 10 (0.2\%) to other European populations, 1 to ``Mexican Ancestry from Los Angeles USA'' (AMR\_MXL) and 1346 (19.9\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
-Among Finns, 1543 (62.4\%) were matched to ``Finnish in Finland'' (EUR\_FIN), 6 (0.3\%) to other European populations, and 922 (37.3\%) could not be matched.
-Among Italians, 229 (22\%) were matched to EUR\_TSI, 15 (1.5\%) to EUR\_IBS and 795 (76.5\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
+We also test the approach proposed in \cite{zhang2020fast} which consists of finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
+When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
+Less than 0.5\% could not be matched by their method (Table \ref{tab:ancestry-pred-kNN}).
+Of note, they could match much more admixed individuals, whereas they set a high probably aiming at discarding such admixed individuals. 
+Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
+Finally, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-fine-pred}).
+We also applied our ancestry detection technique to the European individuals of the POPRES data \cite[]{nelson2008population}. 
+Only 16 out of the 1385 individuals (1.2\%) could not be matched, of which 11 were from East or South-East Europe (Table \ref{tab:ancestry-pred-popres}).
+Note that all individuals that we could match were identified as of European ancestry. 
+We could also identify accurately sub-regions of Europe; e.g.\ 261 out of 264 Spanish and Portugese individuals were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Sun Sep 27 11:26:46 2020
-\begin{table}[ht]
+\begin{table}[htb]
 	\centering
@@
 
-We also test the approach proposed in \cite{zhang2020fast} which consists of finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
-When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[h]
+Finally, we show how to use our ancestry inference method for grouping genetically homogeneous individuals.
+We identify 3 possible ways of doing this.
+The first approach is to simply match identify individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
+We can e.g. define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
+Likewise, we can define ``AFR'', ``AMR'', ``EAS'' and ``SAS'' clusters of individuals who are close enough to any of the corresponding 1000G populations.
+We can then verify that these clusters define distinct sets of homogeneous individuals (Figure \ref{fig:grouping}).
+Another solution, which does not require projecting individuals to the 1000G, but does require computing PC scores in the dataset instead, would consist in using reported ancestries and internal PC scores.
+Then, a British ancestry cluster could be defined by individuals close enough to the center of the people reported as ``British''  (Figure \ref{fig:hist2}).
+We can also define an ``AFR2'' cluster if close enough to the center of either ``Caribbean'', ``African'' or ``Other Black'' individuals, ``EAS2'' if close to the ``Chinese'' cluster, and ``SAS2'' if close to either ``Indian'', ``Pakistani'' or ``Bangladeschi'' clusters.
+This enables to capture a larger set of individuals who are close enough to British people (e.g. Irish people), while discarding individuals whose genetic ancestry is not really ``British''. 
+A third solution, useful when we know that the data is composed of an overdominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
+We recall the we are using the geometric median as such center, which is robust to outliers.
+Overall, all these techniques approximately identify the same amount of people (Table \ref{tab:ancestry-groups}).
+
+\begin{figure}[htb]
 	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
@@
 Contrast with Mahalanobis distance proposed before.
+
+Number of PCs -> 2 is not enough
 
@@
 
-\subsection*{Definitions $\dag$}
+\subsection*{Definitions $\dag$ and methods}
+
+[ALSO DESCRIBE 1000G DATA]
 
@@
 We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
+
+	\item The {\bf UK Biobank} is a large cohort of half a million individuals from the UK, for which we have access to both genotypes and multiple phenotypes (\url{https://www.ukbiobank.ac.uk/}).
+	We apply some quality control filters to the genotyped data; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
+	This results in 488,371 individuals and 504,139 genetic variants.
 
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{proj-UKBB-to-1000G}}
-	\caption{ \label{fig:proj-UKBB}}
+	\caption{First 18 PC scores of the 1000G data (in black), onto which the UK Biobank data has been projected (in red). \label{fig:proj-UKBB}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-mismatch}}
-	\caption{ \label{fig:mismatch}}
+	\caption{PC scores (computed in the UK Biobank) colored by self-reported ancestry.
+On the left, these are 50,000 random individuals.
+On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and or ancestry estimation.
+\label{fig:mismatch}}
 \end{figure}
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
+\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
 \caption{ \label{fig:hist1}}
 \end{figure}
+
 \begin{figure}[h]
-
-\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
+	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
 	\caption{ \label{fig:hist2}}

### 2020-10-05 code/project-UKBB-to-1000G.R

@@ -196,7 +196,7 @@ all_prob <- do.call("rbind", lapply(rows_along(proj2), function(i) {
 ind_ok <- which(all_prob > 0.875, arr.ind = TRUE)
 pred_pop <- rep(NA, nrow(proj2))
 pred_pop[ind_ok[, 1]] <- levels(super_pop)[ind_ok[, 2]]
-table(pred_pop, exclude = NULL)
+table(pred_pop, exclude = NULL)  # 2225 missing
 
 library(magrittr)
 table(pop_UKBB, pred_pop, exclude = NULL) %>%

### 2020-10-05 paper/paper-ancestry-matching.tex
add discu

@@
 Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the minimum distance to their respective center (geometric median$^\dag$).
-Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.
-0.002.
+Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
 This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
@@
 This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries.
-Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g. 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian`.
+Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian`.
 Only 47 individuals were wrongly classified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
@@
 The first approach is to simply match identify individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
-We can e.g. define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
+We can e.g.\ define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
 Likewise, we can define ``AFR'', ``AMR'', ``EAS'' and ``SAS'' clusters of individuals who are close enough to any of the corresponding 1000G populations.
@@
 We can also define an ``AFR2'' cluster if close enough to the center of either ``Caribbean'', ``African'' or ``Other Black'' individuals, ``EAS2'' if close to the ``Chinese'' cluster, and ``SAS2'' if close to either ``Indian'', ``Pakistani'' or ``Bangladeschi'' clusters.
-This enables to capture a larger set of individuals who are close enough to British people (e.g. Irish people), while discarding individuals whose genetic ancestry is not really ``British''. 
-A third solution, useful when we know that the data is composed of an overdominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
+This enables to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British''. 
+A third solution, useful when we know that the data is composed of an predominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
 We recall the we are using the geometric median as such center, which is robust to outliers.
@@
 
-Global measures vs relative measures.
+Here, we propose a PCA-based method for ancestry inference and grouping into genetically homogeneous clusters.
+We show how the PCA-based distance is related to $F_{ST}$, which allows to compute distances based on PC scores directly.
+Previously, we had proposed to use robust Mahalanobis distances to identify a homogeneous group of individuals \cite[]{prive2020efficient}. 
+When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appear suboptimal here compared to using a simple Euclidean distance.
+We hypothesize that the main issue with this approach is that, when there is a very tight cluster, all distances to this cluster are large because of the covariance component in the Mahalanobis distance.
+In contrast, here we propose to directly use the global scale of the PC scores, which is invariant from the cluster scattering.
+This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN).
+Indeed, consider an example of an admixed individual with say 25\% European ancestry and 75\% African ancestry.
+The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested defining genetically homogeneous groups.
 
-Contrast with Mahalanobis distance proposed before.
+Yet, our proposed method has some limitations. 
+First, since we match target individuals to 1000G populations, if individuals are far from all 26 1000G populations, then they would not be matched.
+When looking at the POPRES data, more individuals from East Europe could not be matched. 
+This is not surprising because there are no Eastern European population in the 1000G data.
+Moreover, if we look at the location of the 1000G populations on a map (Figure \ref{fig:map}), we can see that it lacks representation of many parts of the world.
+This issue has been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
+More diverse populations should be aggregated to better cover the worldwide genome diversity.
+This probably would make our method better and, if a sufficient coverage is achieved, we could approximate geographic coordinates by triangulation.
+Then, we also show how to define homogeneous ancestry groups without the use of 1000G data.
+When an predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, it is possible to directly rely on this to restrict to an homogeneous subset of the data.
 
-Number of PCs -> 2 is not enough
+A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster above which the ancestry is not matched.
+Many studies use only the first two PCs for ancestry inference.
+This is not enough to distinguish between Admixed Americans and South Asians (see AMR and SAS in figure \ref{fig:grouping}), let alone for distinguishing between populations at the sub-continental level.
+As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
+Moreover, we believe our method to be robust to the number of PCs used because contribution of later PCs is smaller than for first PCs.
+For the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
+Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
+This threshold can also be adjusted depending on how stringent you want your selection to be.
 
-\noindent[DISCUSS LACK OF SOME POPULATIONS IN 1000G]
-
-\noindent[DISCUSS PC-BASED GENETIC DISTANCES BETWEEN POPULATIONS]
-
-\noindent[IMPLICATIONS FOR PRS?]
+In conclusion, we believe our proposed approach to be an easy way to infer global ancestry and to define groups of homogeneous ancestry.
+This not a method to define admixture coefficients or infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast}.
+However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in ancestry homogeneous samples.
+This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
 
@@
 
-%R package pcadapt is available on CRAN.
-%A tutorial on using pcadapt to detect local adaptation is available at \url{https://bcm-uga.github.io/pcadapt/articles/pcadapt.html}.
-%The code used in this paper is available at \url{https://github.com/bcm-uga/pcadapt/tree/master/new-paper/code}.
+The code used in this paper is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.
 
@@
 \centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
-\caption{ \label{fig:hist1}}
+\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data.
+The threshold in red assumes an $F_{ST}$ of 0.002.
+\label{fig:hist1}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
-	\caption{ \label{fig:hist2}}
+	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals.
+The threshold in red assumes an $F_{ST}$ of 0.002.
+\label{fig:hist2}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
-	\caption{ \label{fig:hist3}}
+	\caption{Histogram of (log) squared distances from the UK Biobank PC scores to the geometric median of the all UKBB individuals.
+Here we use a threshold at 7, based on visual inspection.
+\label{fig:hist3}}
 \end{figure}

### 2020-10-06 paper/paper-ancestry-matching.tex
add intro

@@
 
-Why do we need to group individuals in ancestry groups?
+Principal Component Analysis (PCA) has been widely used to correct for population structure in association studies and has been shown to mirror geography \cite[]{price2006principal,novembre2008genes}.
+Due to its popularity, many methods has been developed for efficiently computing PCs as well as appropriately projecting samples onto a reference PCA space \cite[]{abraham2017flashpca2,zhang2020fast,prive2020efficient}, making these analyses available for ever increasing datasets in human genetics.
+Naturally, PCA has also been used for ancestry inference. 
+However, among all studies that have used PCA for ancestry inference, we have found there was no consensus on the most appropriate methods for inferring ancestry.
+For example, there may be divergences on which distance metric to use and the number of PCs to use to compute these distances.
 
-There is no consensus on the method to use to infer population structure, but particularly on grouping individuals.
-
-Why it makes sense to use PCA-based distances -> correction for population structure, and geographical distance. + Fast
-
-Global ancestry only
+Here, we first compare several distance metrics with the popular $F_{ST}$ statistic between populations.
+We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
+We do not provide a method to define admixture coefficients nor infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
+However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in ancestry homogeneous samples.
+This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
 
@@
 This enables to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British''. 
-A third solution, useful when we know that the data is composed of an predominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
+A third solution, useful when we know that the data is composed of a predominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
 We recall the we are using the geometric median as such center, which is robust to outliers.
@@
 We show how the PCA-based distance is related to $F_{ST}$, which allows to compute distances based on PC scores directly.
-Previously, we had proposed to use robust Mahalanobis distances to identify a homogeneous group of individuals \cite[]{prive2020efficient}. 
+Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
 When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appear suboptimal here compared to using a simple Euclidean distance.
-We hypothesize that the main issue with this approach is that, when there is a very tight cluster, all distances to this cluster are large because of the covariance component in the Mahalanobis distance.
+We hypothesize that the main issue with this approach is that, when there is a very tight population cluster, all distances to this cluster are large because of the covariance component in the Mahalanobis distance.
 In contrast, here we propose to directly use the global scale of the PC scores, which is invariant from the cluster scattering.
-This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN).
+This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
 Indeed, consider an example of an admixed individual with say 25\% European ancestry and 75\% African ancestry.
 The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested defining genetically homogeneous groups.
+We also believe our technique to be more robust to machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are both two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
+In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
 
-Yet, our proposed method has some limitations. 
+Yet, our proposed method also has limitations. 
 First, since we match target individuals to 1000G populations, if individuals are far from all 26 1000G populations, then they would not be matched.
 When looking at the POPRES data, more individuals from East Europe could not be matched. 
-This is not surprising because there are no Eastern European population in the 1000G data.
+This is not surprising because there are no East European population in the 1000G data.
 Moreover, if we look at the location of the 1000G populations on a map (Figure \ref{fig:map}), we can see that it lacks representation of many parts of the world.
 This issue has been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
-More diverse populations should be aggregated to better cover the worldwide genome diversity.
+Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity.
 This probably would make our method better and, if a sufficient coverage is achieved, we could approximate geographic coordinates by triangulation.
-Then, we also show how to define homogeneous ancestry groups without the use of 1000G data.
-When an predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, it is possible to directly rely on this to restrict to an homogeneous subset of the data.
+Nevertheless, we also show how to define homogeneous ancestry groups without the use of the 1000G data.
+When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, it is possible to directly restrict to a homogeneous subset of the data.
 
 A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster above which the ancestry is not matched.
-Many studies use only the first two PCs for ancestry inference.
+Several studies have used only the first two PCs for ancestry inference.
 This is not enough to distinguish between Admixed Americans and South Asians (see AMR and SAS in figure \ref{fig:grouping}), let alone for distinguishing between populations at the sub-continental level.
 As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
-Moreover, we believe our method to be robust to the number of PCs used because contribution of later PCs is smaller than for first PCs.
+Moreover, we believe our method to be robust to the number of PCs used because contribution of later PCs to the distances is smaller than for first PCs.
 For the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
@@
 
-In conclusion, we believe our proposed approach to be an easy way to infer global ancestry and to define groups of homogeneous ancestry.
-This not a method to define admixture coefficients or infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast}.
-However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in ancestry homogeneous samples.
-This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
+In conclusion, we believe our proposed approach to be a simple and robust way to infer global ancestry and to define groups of homogeneous ancestry.
+It is also very fast, inferring ancestry for 488,371 individuals in 20 minutes.
 
@@
 
+The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
 The code used in this paper is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.
 
-\section*{Acknowledgements}
+%\section*{Acknowledgements}
 
@@
 
+\clearpage
+
 \section*{Supplementary Materials}
@@
 \subsection*{Definitions $\dag$ and methods}
-
-[ALSO DESCRIBE 1000G DATA]
 
@@
 
+	\item The {\bf 1000 Genomes Project (1000G)} data is composed of approximately 100 individuals for each of 26 populations worldwide (described at \url{https://www.internationalgenome.org/category/population/}), including 7 African (AFR), 5 East Asian (EAS), 5 South Asian (SAS), 5 European (EUR) and 4 admixed American (AMR) populations.
+	Here we used the transformed data in PLINK format provided in \cite{prive2020efficient}.
+
 	\item The {\bf UK Biobank} is a large cohort of half a million individuals from the UK, for which we have access to both genotypes and multiple phenotypes (\url{https://www.ukbiobank.ac.uk/}).
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{map-1000G}}
-	\caption{ \label{fig:map}}
+	\caption{Locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population} and represented using R package leaflet. 
+	Note that Gujarati Indian from Houston, Texas were manually moved to Gujarat (22.309425, 72.136230), Sri Lankan Tamil from the UK to Sri Lanka (6.927079,	79.861244), and Indian Telugu from the UK to (16.5, 79.5) to better reflect the location of their ancestors.
+	Also note that Utah Residents with Northern and Western European Ancestry, Americans of African Ancestry in SW USA, African Caribbeans in Barbados and Mexican Ancestry from Los Angeles USA are not located at their ancestral location.
+\label{fig:map}}
 \end{figure}

### 2020-10-06 code/geomap-1000G.R

@@ -9,12 +9,12 @@ df <- data.frame(pop  = all_pop$Population,
                  lat  = igsr_pop$`Population latitude`[ind],
                  lon  = igsr_pop$`Population longitude`[ind])
 
-Amsterdam <- c(52.377956, 4.897070)
-df[df$pop == "CEU", c("lat", "lon")] <- Amsterdam
 Gujarat <- c(22.309425, 72.136230)
 df[df$pop == "GIH", c("lat", "lon")] <- Gujarat
 df[df$pop == "ITU", c("lat", "lon")] <- c(16.5, 79.5)
 df[df$pop == "STU", c("lat", "lon")] <- c(6.927079,	79.861244)
+# Amsterdam <- c(52.377956, 4.897070)
+# df[df$pop == "CEU", c("lat", "lon")] <- Amsterdam
 # Edinburgh <- c(55.953251, -3.188267)
 # London <- c(51.509865, -0.118092)
 # df[df$pop == "GBR", c("lat", "lon")] <- (Edinburgh + London) / 2

### 2020-10-06 paper/paper-ancestry-matching.tex
first draft finished

@@
 
-$^\dag$ Further defined in supplementary section ``Defintions''.
-
 \subsection*{Introduction}
@@
 Naturally, PCA has also been used for ancestry inference. 
-However, among all studies that have used PCA for ancestry inference, we have found there was no consensus on the most appropriate methods for inferring ancestry.
+However, among all studies where we have seen PCA used for ancestry inference, we have found there was no consensus on the most appropriate methods for inferring ancestry.
 For example, there may be divergences on which distance metric to use and the number of PCs to use to compute these distances.
@@
 We do not provide a method to define admixture coefficients nor infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
-However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in ancestry homogeneous samples.
+However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of ancestry homogeneous.
 This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
+
+\vspace*{2em}
+$^\dag$ Further defined in supplementary section ``Defintions''.
 
@@
 
-We first compare four measures of genetic dissimilarity using populations of the 1000 genomes as example \cite[]{10002015global}. 
+We first compare four measures of genetic dissimilarity using populations of the 1000 Genomes Project (1000G$^\dag$, \cite{10002015global}). 
 The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
-We report $F_{ST}$ between the 26 1000G populations in tables S1-S5.
+We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}.
 The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
@@
 In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1} and \ref{fig:heatmap3}).
-For example, using , African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
+For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
 Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.
@@
 
-We project the dataset of interest (either UKBB or POPRES) onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
+We project the dataset of interest onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
 We recall that this uses a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}. 
-Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the minimum distance to their respective center (geometric median$^\dag$).
+Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the distance to these reference population centers (geometric medians$^\dag$).
 Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
 This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
-When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown.
+When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest population.
 
@@
 Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
-Based on the threshold defined before, we could not infer ancestry for 4.6\% of the individuals.
+Based on the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
 More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}).
 This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries.
-Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian`.
+Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian` ancestries.
 Only 47 individuals were wrongly classified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
-However, when comparing the location of these individuals on the PCA space computed in the UK Biobank \cite[]{bycroft2017genome} to the rest of individuals, it seemed more probable that the self-reported ancestry is wrong and not our ancestry estimate (Figure \ref{fig:mismatch}).
+However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed on the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is wrong (Figure \ref{fig:mismatch}).
 
@@
 
-Finally, we show how to use our ancestry inference method for grouping genetically homogeneous individuals.
-We identify 3 possible ways of doing this.
-The first approach is to simply match identify individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
+Finally, we show 3 possible ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
+The first approach is to simply match individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
 We can e.g.\ define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
@@
 We can then verify that these clusters define distinct sets of homogeneous individuals (Figure \ref{fig:grouping}).
-Another solution, which does not require projecting individuals to the 1000G, but does require computing PC scores in the dataset instead, would consist in using reported ancestries and internal PC scores.
+Another solution, which does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead, would consist in using reported ancestries and internal PC scores.
 Then, a British ancestry cluster could be defined by individuals close enough to the center of the people reported as ``British''  (Figure \ref{fig:hist2}).
 We can also define an ``AFR2'' cluster if close enough to the center of either ``Caribbean'', ``African'' or ``Other Black'' individuals, ``EAS2'' if close to the ``Chinese'' cluster, and ``SAS2'' if close to either ``Indian'', ``Pakistani'' or ``Bangladeschi'' clusters.
-This enables to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British''. 
-A third solution, useful when we know that the data is composed of a predominant ancestry is to keep individuals who are close enough to the center of all individuals (Figure \ref{fig:hist3}).
+This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British'' as self-reported.
+A third solution, useful when we know that the data is composed of a predominant ancestry is to keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
 We recall the we are using the geometric median as such center, which is robust to outliers.
@@
 
-Here, we propose a PCA-based method for ancestry inference and grouping into genetically homogeneous clusters.
-We show how the PCA-based distance is related to $F_{ST}$, which allows to compute distances based on PC scores directly.
+Here we propose a PCA-based method for ancestry inference and grouping into genetically homogeneous clusters.
+We show how the PCA-based distance is related to the $F_{ST}$, which allows to compute distances based on PC scores directly.
 Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
@@
 Indeed, consider an example of an admixed individual with say 25\% European ancestry and 75\% African ancestry.
-The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested defining genetically homogeneous groups.
-We also believe our technique to be more robust to machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are both two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
+The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
+We also believe our proposed method to be more robust than machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
 In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
@@
 Moreover, if we look at the location of the 1000G populations on a map (Figure \ref{fig:map}), we can see that it lacks representation of many parts of the world.
-This issue has been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
+This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
 Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity. 
-This probably would make our method better and, if a sufficient coverage is achieved, we could approximate geographic coordinates by triangulation.
+This probably would make our method better and, if a sufficient coverage is achieved, allow for approximating geographic coordinates by triangulation.
 Nevertheless, we also show how to define homogeneous ancestry groups without the use of the 1000G data.
@@
 As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
-Moreover, we believe our method to be robust to the number of PCs used because contribution of later PCs to the distances is smaller than for first PCs.
-For the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
+Moreover, we believe our proposed method to be robust to the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
+As for the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
 Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
-This threshold can also be adjusted depending on how stringent you want your selection to be.
+This threshold can also be adjusted depending on how stringent you want the selection to be.
 
 In conclusion, we believe our proposed approach to be a simple and robust way to infer global ancestry and to define groups of homogeneous ancestry.
-It is also very fast, inferring ancestry for 488,371 individuals in 20 minutes.
+It is also very fast, allowing to infer ancestry for 488,371 individuals in 20 minutes.
 
@@
 
-%\section*{Acknowledgements}
+\section*{Acknowledgements}
+
+This research has been conducted using the UK Biobank Resource under Application Number 41181.
 
@@
 
+	\item The {\bf 1000 Genomes Project (1000G)} data is composed of approximately 100 individuals for each of 26 populations worldwide (described at \url{https://www.internationalgenome.org/category/population/}), including 7 African (AFR), 5 East Asian (EAS), 5 South Asian (SAS), 5 European (EUR) and 4 admixed American (AMR) populations.
+	Here we used the transformed data in PLINK format provided in \cite{prive2020efficient}.
+
 	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
@@
 
-	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix.
+	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix \cite[]{prive2020efficient}.
 They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.
 
-	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$, which is defined as
+	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ is defined as
 	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
@@
 We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
-
-	\item The {\bf 1000 Genomes Project (1000G)} data is composed of approximately 100 individuals for each of 26 populations worldwide (described at \url{https://www.internationalgenome.org/category/population/}), including 7 African (AFR), 5 East Asian (EAS), 5 South Asian (SAS), 5 European (EUR) and 4 admixed American (AMR) populations.
-	Here we used the transformed data in PLINK format provided in \cite{prive2020efficient}.
 
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{compare-dist-to-Fst}}
-	\caption{Comparing Fst to different distances in the PCA space between pairs of the 26 1000G populations. \label{fig:compare-dist}}
+	\caption{Comparing $F_{ST}$ to different PCA-based distances between pairs of the 26 1000G populations. \label{fig:compare-dist}}
 \end{figure}
@@
 \centerline{\includegraphics[width=0.8\textwidth]{heatmap-Fst-1000G}}
-\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in table S1-S5. \label{fig:heatmap0}}
+\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}. \label{fig:heatmap0}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-closest-1000G}}
-	\caption{Heatmap with clustering based on the shortest distances between pairs of the 26 1000G populations. \label{fig:heatmap3}}
+	\caption{Heatmap with clustering based on the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:heatmap3}}
 \end{figure}
@@
 On the left, these are 50,000 random individuals.
-On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and or ancestry estimation.
+On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and our ancestry estimation.
 \label{fig:mismatch}}

### 2020-10-06 paper/paper-ancestry-matching.tex
add abstract

@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 \abstract{
+	First, we show how Euclidean distance in the Principal Component Analysis (PCA) space is related to $F_{ST}$ between populations.
+	Then, we project the genetic data of individuals from the UK Biobank and the POPRES datasets onto the reference populations of the 1000 Genomes Project and show how to accurately infer global ancestry and group individuals in homogeneous clusters.
+	Inferring ancestry for nearly 500,000 individuals is performed in 20 minutes only.
+	The proposed method is particularly important for methods that needs to be applied in homogeneous samples.
+	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
 }
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
-	\caption{ \label{fig:grouping}}
+	\caption{The UK Biobank PC scores (first four) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
 \end{figure}
@@
 In conclusion, we believe our proposed approach to be a simple and robust way to infer global ancestry and to define groups of homogeneous ancestry.
-It is also very fast, allowing to infer ancestry for 488,371 individuals in 20 minutes.
+It is also very fast, allowing to infer ancestry for 488,371 individuals in 20 minutes using 16 cores.
 

### 2020-10-18 paper/paper-ancestry-matching.tex
save

@@
 	Inferring ancestry for nearly 500,000 individuals is performed in 20 minutes only.
-	The proposed method is particularly important for methods that needs to be applied in homogeneous samples.
+	The proposed method is particularly important for methods that need to be applied in homogeneous samples.
 	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
@@
 We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
-We do not provide a method to define admixture coefficients nor infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
-However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of ancestry homogeneous.
+We do not provide a method to infer admixture coefficients nor local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
+However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of homogeneous ancestry.
 This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
@@
 The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
-We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}.
-The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
-The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
-In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1} and \ref{fig:heatmap3}).
+We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}, and the clustering of these populations based on $F_{ST}$ in figure \ref{fig:heatmap0}.
+The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
+The distance between population centers appears to be an appropriate PCA-based distance as this distance (squared) is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist0}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
+In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:compare-dist}, \ref{fig:heatmap1} and \ref{fig:heatmap3}).
 For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
 Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.
+Note that we used 19 PCs for computing all PCA-based distances based on visual inspection of PC scores, as recommended in \cite{prive2020efficient}. 
+With 2 to 4 PCs, we are able to adequately separate distant populations, but not the closest ones.
+For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are at the same position (Figure \ref{fig:compare-distK}).
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
+	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist0}}
+\end{figure}
 
@@
 This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
-When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest population.
+When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest reference population center.
 
@@
 Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian` ancestries.
-Only 47 individuals were wrongly classified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
-However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed on the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is wrong (Figure \ref{fig:mismatch}).
+Only 47 individuals were misclassified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
+However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).
 
-We also test the approach proposed in \cite{zhang2020fast} which consists of finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
+We also test the approach proposed in \cite{zhang2020fast} which consists in finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
 When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
@@
 
-Finally, we show 3 possible ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
-The first approach is to simply match individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
-We can e.g.\ define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
-Likewise, we can define ``AFR'', ``AMR'', ``EAS'' and ``SAS'' clusters of individuals who are close enough to any of the corresponding 1000G populations.
-We can then verify that these clusters define distinct sets of homogeneous individuals (Figure \ref{fig:grouping}).
-Another solution, which does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead, would consist in using reported ancestries and internal PC scores.
-Then, a British ancestry cluster could be defined by individuals close enough to the center of the people reported as ``British''  (Figure \ref{fig:hist2}).
-We can also define an ``AFR2'' cluster if close enough to the center of either ``Caribbean'', ``African'' or ``Other Black'' individuals, ``EAS2'' if close to the ``Chinese'' cluster, and ``SAS2'' if close to either ``Indian'', ``Pakistani'' or ``Bangladeschi'' clusters.
-This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British'' as self-reported.
-A third solution, useful when we know that the data is composed of a predominant ancestry is to keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
-We recall the we are using the geometric median as such center, which is robust to outliers.
-Overall, all these techniques approximately identify the same amount of people (Table \ref{tab:ancestry-groups}).
+Finally, we show several ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
+The first approach is to simply match individuals that are close enough to one of the 1000G populations, as described previously.
+
+Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth.
+This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.
+We can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all of them).
+This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (TODO: TABLE).
+The resulting clusters are presented on the PC space in figure \ref{fig:grouping}.
+
+One could do the same using the countries of birth (Field 20115) instead of the self-reported ancestries.
+Again, the country of birth may sometime not reflect the ancestral origin.
+Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals.
+Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset with good coverage of the overall dissimilarities (TODO FIGURE HEATMAP).
+Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
+Interestingly, the individuals in the clusters from Iran and Poland were not previously identified when using the self-reported ancestries, while some individuals previously identified as ``Bangladeshi'' are not matched to India (TODO FIGURES).
+2.8\% of individuals could not be matched using countries of birth compared to 3.7\% when using the self-reported ancestry (TODO TABLES).
+
+Finally, when we know that the data is composed of a predominant ancestry, we can simply keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
+When doing so, we can cluster 91\% of the data into one cluster composed of 421,871 British, 12,039 Irish, 8351 ``Other White'', 1814 individuals of unknown ancestry, 467 ``White'' and 41 individuals of other self-reported ancestries.
+This is made possible because we use the geometric median which is robust to outliers.
 
@@
 
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
+	\caption{
+		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
+Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
+		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
+		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
+		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
+\label{fig:map}}
+\end{figure}
+
 A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster above which the ancestry is not matched.
@@
 
+Authors thank Alex Diaz-Papkovich, Clive Hoggart and others for their useful feedback.
 This research has been conducted using the UK Biobank Resource under Application Number 41181.
@@
 
-\subsection*{Location of 1000G populations}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{map-1000G}}
-	\caption{Locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population} and represented using R package leaflet. 
-	Note that Gujarati Indian from Houston, Texas were manually moved to Gujarat (22.309425, 72.136230), Sri Lankan Tamil from the UK to Sri Lanka (6.927079,	79.861244), and Indian Telugu from the UK to (16.5, 79.5) to better reflect the location of their ancestors.
-	Also note that Utah Residents with Northern and Western European Ancestry, Americans of African Ancestry in SW USA, African Caribbeans in Barbados and Mexican Ancestry from Los Angeles USA are not located at their ancestral location.
-\label{fig:map}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
 \subsection*{Measures of genetic dissimilarity between populations}
@@
 	\caption{Comparing $F_{ST}$ to different PCA-based distances between pairs of the 26 1000G populations. \label{fig:compare-dist}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
+	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations.
+Distances are computed using a different number of Principal Components (PCs) \label{fig:compare-distK}}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
-\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data.
+%\begin{figure}[h]
+%\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
+%\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data.
 The threshold in red assumes an $F_{ST}$ of 0.002.
 \label{fig:hist1}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
-	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals.
+%\end{figure}
+%
+%\begin{figure}[h]
+%	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
+%	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals.
 The threshold in red assumes an $F_{ST}$ of 0.002.
 \label{fig:hist2}}
-\end{figure}
+%\end{figure}
 
@@
 Here we use a threshold at 7, based on visual inspection.
+Alternatively, a more stringent threshold at 6 could also be used.
 \label{fig:hist3}}

### 2020-10-25 code/ancestry-grouping-without-1000G.R
reexport code

@@ -1,0 +1,233 @@
+library(bigsnpr)
+library(dplyr)
+library(ggplot2)
+
+print_table <- function(x) {
+  tab <- sort(table(x, exclude = NULL), decreasing = TRUE)
+  cat(paste0(names(tab), ": ", tab), sep = " || ")
+}
+
+#### Groupings of homogeneous ancestry  ####
+
+df <- readRDS("data/info_UKBB.rds")
+pop_UKBB <- df$pop
+PC_UKBB <- as.matrix(select(df, PC1:PC16))
+
+## 1) No known ancestry (just knowing there is an overdominant population)
+
+round(one_center <- bigutilsr::geometric_median(na.omit(PC_UKBB)), 4)
+#      PC1      PC2      PC3      PC4      PC5      PC6      PC7      PC8
+# -11.6409   3.4545  -1.3283   0.7628  -1.2801  -0.2577   0.2160  -0.3044
+#      PC9     PC10     PC11     PC12     PC13     PC14     PC15     PC16
+#   0.2386   0.0030  -0.2657   0.1783   0.0167  -0.0744  -0.0025  -0.1004
+round(bigutilsr::covrob_ogk(na.omit(PC_UKBB))$center, 4)
+#      PC1      PC2      PC3      PC4      PC5      PC6      PC7      PC8
+# -12.3377   3.7848  -1.5816   1.2128  -1.1087  -0.3453   0.2632  -0.3986
+#      PC9     PC10     PC11     PC12     PC13     PC14     PC15     PC16
+#   0.6842  -0.0162  -0.3409   0.2494   0.0271   0.0214  -0.0155  -0.1531
+
+all_sq_dist <- bigutilsr:::rowSumsSq(sweep(PC_UKBB, 2, one_center, '-'))
+
+qplot(log(all_sq_dist)) +
+  theme_bigstatsr(0.9) +
+  labs(x = "Log squared distance to overall center") +
+  scale_x_continuous(breaks = 0:20)
+# ggsave("figures/hist-dist-overall-center.pdf", width = 7, height = 5)
+
+thr_sq_dist <- exp(7)  # visual inspection of histogram
+
+mean(all_sq_dist < thr_sq_dist, na.rm = TRUE) # 91.06%
+ind_homogeneous <- which(all_sq_dist < thr_sq_dist)
+
+print_table(as.character(pop_UKBB[ind_homogeneous]))
+# British: 421871 || Irish: 12039 || Other White: 8351 || NA: 1814 ||
[…]

### 2020-10-25 code/ancestry-grouping.R

@@ -1,225 +1,0 @@
-library(bigsnpr)
-library(ggplot2)
-NCORES <- 15
-
-#### Groupings of homogeneous ancestry  ####
-
-## 1) Matching to 1000G populations
-
-# Project UKBB individuals onto PCA space from 1000G
-
-obj.bed <- bed("data/ukbb.bed")
-bed.ref <- bed(download_1000G("tmp-data"))
-ind.norel <- readRDS("tmp-data/ind-norel-1000G.rds")
-
-proj <- runonce::save_run(
-  bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.norel,
-                 k = 25, ncores = NCORES),
-  file = "data/proj-UKBB-to-1000G.rds"
-) # < 15 min
-# 1,664,852 variants to be matched.
-# 22,794 ambiguous SNPs have been removed.
-# 377,914 variants have been matched; 0 were flipped and 65,346 were reversed.
-
-PC.ref <- predict(proj$obj.svd.ref)
-proj2 <- proj$OADP_proj
-
-fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")[ind.norel, ]
-pop_1000G <- paste0(fam2$`Super Population`, "_", fam2$Population)
-
-seq_PC <- 1:16
-thr_Fst <- 0.002
-
-all_centers <- bigutilsr::geometric_median(PC.ref[, seq_PC], by_grp = pop_1000G)
-thr_sq_dist <- thr_Fst * (max(dist(all_centers)^2) / 0.16)
-
-all_sq_dist <- apply(all_centers, 1, function(center) {
-  bigutilsr:::rowSumsSq(sweep(proj2[, seq_PC], 2, center, '-'))
-})
-
-## British ancestry (and around)
-qplot(all_sq_dist[, "EUR_GBR"]) +
-  theme_bigstatsr(0.9) +
-  scale_x_log10() +
-  labs(x = "Squared distance to GBR cluster") +
-  geom_vline(xintercept = thr_sq_dist, color = "red")
-# ggsave("figures/hist-dist-GBR.pdf", width = 7, height = 5)
-
-mean(all_sq_dist[, "EUR_GBR"] < thr_sq_dist) # 89.73%
-ind_GBR <- which(all_sq_dist[, "EUR_GBR"] < thr_sq_dist)
-
-df <- readRDS("data/info_UKBB.rds")
[…]

### 2020-10-25 code/dist-pop-1000G.R

@@ -1,4 +1,6 @@
 library(bigsnpr)
+library(ggplot2)
+theme_set(theme_bigstatsr())
 NCORES <- 15
 
 #### Compute PCA on 1000G ####
@@ -25,7 +27,7 @@ obj.svd <- runonce::save_run(
   file = "tmp-data/SVD-1000G.rds"
 )
 plot(obj.svd, type = "scores", scores = 14:25, coeff = 0.6)
-PC <- predict(obj.svd)[, 1:19]
+PC <- predict(obj.svd)
 
 
 #### Compute multiple distances between 1000G populations ####
@@ -102,6 +104,12 @@ pdf("figures/heatmap-bhattacharyya-1000G.pdf", width = 8, height = 8)
 heatmap(dist_bhattacharyya, symm = TRUE, Rowv = as.dendrogram(hc))
 dev.off()
 
+qplot(dist_bhattacharyya, dist_fst) +
+  scale_color_viridis_c() +
+  geom_smooth(color = "red", method = "lm", formula = "y ~ x + 0") +
+  labs(x = "Bhattacharyya distance between clusters", y = "Fst")
+# ggsave("figures/compare-Bhattacharyya-to-Fst.pdf", width = 8, height = 6)
+
 
 ## Distance between centers
 
@@ -112,6 +120,25 @@ hc <- hclust(dist_centers, method = "single")
 pdf("figures/heatmap-centers-1000G.pdf", width = 8, height = 8)
 heatmap(as.matrix(dist_centers), symm = TRUE, Rowv = as.dendrogram(hc))
 dev.off()
+
+qplot(as.matrix(dist_centers)^2, dist_fst) +
+  scale_color_viridis_c() +
+  geom_smooth(color = "red", method = "lm", formula = "y ~ x + 0") +
+  labs(x = "Squared distance between centers of clusters", y = "Fst")
+# ggsave("figures/compare-Euclidean-to-Fst.pdf", width = 8, height = 6)
+
+all_plots <- lapply(c(2, 3, 4, 8, 16, 25), function(K) {
+  all_centers <- bigutilsr::geometric_median(PC[, 1:K], by_grp = pop_1000G)
+  dist_centers <- dist(all_centers)
+  qplot(as.matrix(dist_centers)^2, dist_fst) +
+    scale_color_viridis_c() +
+    theme_bigstatsr(0.7) +
+    geom_smooth(color = "red", method = "lm", formula = "y ~ x + 0") +
[…]

### 2020-10-25 code/plot-unmatched.R

@@ -1,0 +1,89 @@
+df <- readRDS("data/matched-pop.rds")
+table(df$continent)
+#   Africa      Asia    Europe   North America    Oceania   South America
+#    12244     11421    445233            2618       1663            1327
+
+library(dplyr)
+library(ggplot2)
+
+# Make sure country names are matching
+df2 <- df %>%
+  mutate(country = forcats::fct_recode(
+    country,
+    "Democratic Republic of the Congo" = "Congo",
+    "Republic of Congo" = "Congo",
+    "Burkina Faso" = "Burkina",
+    "Madeira Islands" = "Madeira",
+    "United Arab Emirates" = "Emirates",
+    "Serbia/Montenegro" = "Serbia",
+    "China" = "Hong Kong"
+  ))
+
+world_map <- map_data("world") %>%
+  mutate(region = forcats::fct_recode(
+    region,
+    "United Kingdom" = "UK",
+    "South Georgia and the South Sandwich Islands" = "South Sandwich Islands",
+    "South Georgia and the South Sandwich Islands" = "South Georgia",
+    "Myanmar (Burma)" = "Myanmar",
+    "Republic of Kosovo" = "Kosovo",
+    "Antigua and Barbuda" = "Antigua",
+    "Antigua and Barbuda" = "Barbuda",
+    "Serbia/Montenegro" = "Serbia"))
+
+all_countries <- unique(df2$country)
+world_regions <- unique(world_map$region)
+not_found <- all_countries[!all_countries %in% world_regions]
+table(df2$country)[not_found]
+sapply(not_found, grep, x = world_regions, value = TRUE) %>%
+  setNames(not_found) %>%
+  { .[lengths(.) != 0] }
+sapply(world_regions, grep, x = not_found, value = TRUE) %>%
+  setNames(world_regions) %>%
+  { .[lengths(.) != 0] }
+
+
+p <- ggplot({
+  df2 %>%
+    group_by(country) %>%
+    summarise(unmatched = if (n() > 30) 100 * mean(is.na(infer)) else NA) %>%
+    arrange(desc(unmatched)) %>%
+    print(n = Inf) %>%
+    right_join(world_map, by = c("country" = "region"))
+}) +
+  bigstatsr::theme_bigstatsr() +
+  # coord_equal() +
+  geom_polygon(aes(long, lat, group = group, fill = unmatched),
[…]

### 2020-10-25 code/prepare-UKBB.R

@@ -29,11 +29,13 @@ library(dplyr)
 code_ancestry <- fread2("UKBB/coding1001.tsv")
 unknown <- c("Prefer not to answer", "Do not know", "Mixed",
              "Other ethnic group", "Any other mixed background")
+code_country <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+code_continent <- filter(fread2("UKBB/coding89.tsv"), selectable == "N")
 
 df0 <- fread2(
   "UKBB/ukb41181.csv",
-  select = c("eid", "21000-0.0", paste0("22009-0.", 1:16)),
-  col.names = c("eid", "pop", paste0("PC", 1:16))
+  select = c("eid", "21000-0.0", "20115-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "pop", "country", paste0("PC", 1:16))
 ) %>%
   mutate(
     pop  = factor(pop, levels = code_ancestry$coding,
@@ -46,9 +48,19 @@ df0 <- fread2(
       forcats::fct_relevel(c(
         "British", "Irish", "White", "Other White",
         "Indian", "Pakistani", "Bangladeshi", "Chinese", "Other Asian",
-        "Caribbean", "African", "Other Black"))
+        "Caribbean", "African", "Other Black")),
+
+    continent = factor(country, levels = code_country$coding,
+                       labels = code_country$parent_id) %>%
+      factor(labels = code_continent$meaning),
+
+    country = factor(country, levels = code_country$coding,
+                     labels = code_country$meaning)
   )
 str(df0)
+df0$country[with(df0, is.na(country) & pop == "British")] <- "United Kingdom"
+df0$country[with(df0, is.na(country) & pop == "Irish")]   <- "Ireland"
+df0$continent[df0$country %in% c("United Kingdom", "Ireland")] <- "Europe"
 
 bed_eid <- fread2("data/ukbb.fam")[[2]]
 df <- df0[match(bed_eid, df0$eid), ]

### 2020-10-25 code/project-UKBB-to-1000G.R

@@ -73,6 +73,10 @@ mean(is.na(choose_pop2)) # 4.55%
 
 df <- readRDS("data/info_UKBB.rds")
 pop_UKBB <- df$pop
+saveRDS(data.frame(pop = pop_UKBB, infer = choose_pop2,
+                   continent = df$continent, country = df$country),
+        "data/matched-pop.rds")
+
 mixed <- c("Asian or Asian British", "Black or Black British",
            "White and Black Caribbean", "White and Black African",
            "White and Asian")

### 2020-10-25 paper/paper-ancestry-matching.tex
some changes to text

@@
 \abstract{
-	First, we show how Euclidean distance in the Principal Component Analysis (PCA) space is related to $F_{ST}$ between populations.
-	Then, we project the genetic data of individuals from the UK Biobank and the POPRES datasets onto the reference populations of the 1000 Genomes Project and show how to accurately infer global ancestry and group individuals in homogeneous clusters.
-	Inferring ancestry for nearly 500,000 individuals is performed in 20 minutes only.
-	The proposed method is particularly important for methods that need to be applied in homogeneous samples.
+	Here, we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
+	The proposed approach is particularly useful for methods that need to be applied in homogeneous samples.
+	First, we show that Euclidean distances in the PCA space are proportional to $F_{ST}$ between populations.
+	Then, we show how to use this PCA-based distance to infer ancestry in the UK Biobank and the POPRES datasets.
+	We propose two solutions, either relying on projection of PCs to reference populations such as from the 1000 Genomes Project, or by directly using the internal data.
 	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
@@
 
-Principal Component Analysis (PCA) has been widely used to correct for population structure in association studies and has been shown to mirror geography \cite[]{price2006principal,novembre2008genes}.
-Due to its popularity, many methods has been developed for efficiently computing PCs as well as appropriately projecting samples onto a reference PCA space \cite[]{abraham2017flashpca2,zhang2020fast,prive2020efficient}, making these analyses available for ever increasing datasets in human genetics.
+Principal Component Analysis (PCA) has been widely used to correct for population structure in association studies and has been shown to mirror geography in Europe \cite[]{price2006principal,novembre2008genes}.
+Due to its popularity, many methods has been developed for efficiently performing PCA \cite[]{abraham2017flashpca2,prive2020efficient} as well as appropriately projecting samples onto a reference PCA space \cite[]{zhang2020fast,prive2020efficient}, making possible to perform these analyses for ever increasing datasets in human genetics.
 Naturally, PCA has also been used for ancestry inference. 
-However, among all studies where we have seen PCA used for ancestry inference, we have found there was no consensus on the most appropriate methods for inferring ancestry.
+However, among all studies where we have seen PCA used for ancestry inference, we have found there was no consensus on what is the most appropriate method for inferring ancestry using PCA.
 For example, there may be divergences on which distance metric to use and the number of PCs to use to compute these distances.
@@
 However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of homogeneous ancestry.
-This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.
+This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the population used for training \cite[]{martin2017human}.
 
@@
 We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}, and the clustering of these populations based on $F_{ST}$ in figure \ref{fig:heatmap0}.
-The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
-The distance between population centers appears to be an appropriate PCA-based distance as this distance (squared) is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist0}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
-In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:compare-dist}, \ref{fig:heatmap1} and \ref{fig:heatmap3}).
-For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
-Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.
-Note that we used 19 PCs for computing all PCA-based distances based on visual inspection of PC scores, as recommended in \cite{prive2020efficient}. 
+The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric medians$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
+The (squared) distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
+In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1}, \ref{fig:compare-dist1}, \ref{fig:heatmap3} and \ref{fig:compare-dist3}).
+For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) and the four admixed American (AMR) populations are close to all European (EUR), South Asian (SAS) and African (AFR) populations when using the Bhattacharyya distance (Figure \ref{fig:heatmap1}).
+Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers.
+Note that we use 16 PCs for computing all PCA-based distances based on visual inspection of PC scores, as recommended in \cite{prive2020efficient}. 
 With 2 to 4 PCs, we are able to adequately separate distant populations, but not the closest ones.
-For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are at the same position (Figure \ref{fig:compare-distK}).
+For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are superimposed (Figure \ref{fig:compare-distK}).
 
@@
 	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist0}}
+	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
 \end{figure}
@@
 We project the dataset of interest onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
-We recall that this uses a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}. 
-Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the distance to these reference population centers (geometric medians$^\dag$).
+We recall that this uses an automatic removal of LD when computing PCA and a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}.
+Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers (geometric medians$^\dag$) in the PCA space.
 Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
@@
 We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
-These individuals seem to originate from all over the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
+These individuals seem to originate from many parts of the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
 Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
@@
 This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries.
-Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian` ancestries.
+Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within ``White and Black Africa'' and 93.0\% within ``White and Asian'' ancestries.
 Only 47 individuals were misclassified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
 However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).
+This possible discrepancy between self-reported ancestry and genetic ancestry has been reported before \cite[]{mersha2015self}.
 
@@
 Less than 0.5\% could not be matched by their method (Table \ref{tab:ancestry-pred-kNN}).
-Of note, they could match much more admixed individuals, whereas they set a high probably aiming at discarding such admixed individuals. 
+Of note, they could match much more admixed individuals, whereas they set a high probability threshold aiming at discarding such admixed individuals. 
 Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
@@
 
-Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth.
+Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth, e.g.\ available in the UK Biobank (Fields 21000 and 20115). 
 This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.
-We can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all of them).
+In the UK Biobank data, we can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all centers).
 This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (TODO: TABLE).
-The resulting clusters are presented on the PC space in figure \ref{fig:grouping}.
+Only 3.7\% of all individuals could not be matched.
+The resulting clusters are presented on the PCA space in figure \ref{fig:grouping}.
 
-One could do the same using the countries of birth (Field 20115) instead of the self-reported ancestries.
+One could do the same using the countries of birth instead of the self-reported ancestries.
 Again, the country of birth may sometime not reflect the ancestral origin.
 Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals.
-Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset with good coverage of the overall dissimilarities (TODO FIGURE HEATMAP).
+Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
 Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
-Interestingly, the individuals in the clusters from Iran and Poland were not previously identified when using the self-reported ancestries, while some individuals previously identified as ``Bangladeshi'' are not matched to India (TODO FIGURES).
-2.8\% of individuals could not be matched using countries of birth compared to 3.7\% when using the self-reported ancestry (TODO TABLES).
+Only 2.8\% of all individuals could not be matched (TODO TABLE).
+The resulting clusters are presented on the PCA space in figure \ref{fig:grouping2}.
+Note that these clusters probably include individuals from nearby countries as well.
 
-Finally, when we know that the data is composed of a predominant ancestry, we can simply keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
+Finally, when we know that the data is composed of a predominant ancestry, we can define a single homogeneous cluster by simply restricting to individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
 When doing so, we can cluster 91\% of the data into one cluster composed of 421,871 British, 12,039 Irish, 8351 ``Other White'', 1814 individuals of unknown ancestry, 467 ``White'' and 41 individuals of other self-reported ancestries.
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
-	\caption{The UK Biobank PC scores (first four) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
+	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
 \end{figure}
@@
 
-Here we propose a PCA-based method for ancestry inference and grouping into genetically homogeneous clusters.
+Here we propose a PCA-based method for ancestry inference and grouping individuals into genetically homogeneous clusters.
 We show how the PCA-based distance is related to the $F_{ST}$, which allows to compute distances based on PC scores directly.
-Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
-When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appear suboptimal here compared to using a simple Euclidean distance.
-We hypothesize that the main issue with this approach is that, when there is a very tight population cluster, all distances to this cluster are large because of the covariance component in the Mahalanobis distance.
+This relation between $F_{ST}$ and (squared) Euclidean distances in the PCA space has been previously shown for two populations only \cite[]{mcvean2009genealogical}.
+Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a single homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
+When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appears suboptimal here compared to using a simple Euclidean distance.
+We hypothesize that the main issue with this approach is that an admixed population covers a large volume in the PCA space, therefore all distances to this population cluster are small because of the covariance component from the Mahalanobis distance.
 In contrast, here we propose to directly use the global scale of the PC scores, which is invariant from the cluster scattering.
 This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
-Indeed, consider an example of an admixed individual with say 25\% European ancestry and 75\% African ancestry.
+Indeed, consider an example of an admixed individual of say 25\% European ancestry and 75\% African ancestry.
 The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
@@
 This is not surprising because there are no East European population in the 1000G data.
-Moreover, if we look at the location of the 1000G populations on a map (Figure \ref{fig:map}), we can see that it lacks representation of many parts of the world.
+Moreover, if we look at the location of the 1000G populations on a map, we can see that it lacks representation of many parts of the world (Figure \ref{fig:map}).
 This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
-Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity. 
-This probably would make our method better and, if a sufficient coverage is achieved, allow for approximating geographic coordinates by triangulation.
-Nevertheless, we also show how to define homogeneous ancestry groups without the use of the 1000G data.
+Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve our proposed method.
+Nevertheless, we also show how to define homogeneous ancestry groups without using the 1000G data.
 When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, it is possible to directly restrict to a homogeneous subset of the data.
@@
 
-A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster above which the ancestry is not matched.
+A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.
 Several studies have used only the first two PCs for ancestry inference.
-This is not enough to distinguish between Admixed Americans and South Asians (see AMR and SAS in figure \ref{fig:grouping}), let alone for distinguishing between populations at the sub-continental level.
+This is not enough to distinguish between Admixed Americans and South Asians (see AMR and SAS in figure \ref{fig:grouping}), let alone for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
 As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
@@
 Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
-This threshold can also be adjusted depending on how stringent you want the selection to be.
+This threshold can also be adjusted depending on how homogeneous one want each cluster to be.
 
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-dist-to-Fst}}
-	\caption{Comparing $F_{ST}$ to different PCA-based distances between pairs of the 26 1000G populations. \label{fig:compare-dist}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
-	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations.
-Distances are computed using a different number of Principal Components (PCs) \label{fig:compare-distK}}
-\end{figure}
+\FloatBarrier
 
@@
 \begin{figure}[h]
+	\centerline{\includegraphics[width=0.7\textwidth]{compare-Bhattacharyya-to-Fst}}
+	\caption{Comparing $F_{ST}$ to the Bhattacharyya distance on the PCA space between pairs of the 26 1000G populations. \label{fig:compare-dist1}}
+\end{figure}
+
+\begin{figure}[h]
 	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-centers-1000G}}
@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
+\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations.
+Distances are computed using different numbers of Principal Components (PCs).
+\label{fig:compare-distK}}
+\end{figure}
+
+\begin{figure}[h]
 	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-closest-1000G}}
 	\caption{Heatmap with clustering based on the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:heatmap3}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.7\textwidth]{compare-closest-to-Fst}}
+	\caption{Comparing $F_{ST}$ to the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:compare-dist3}}
 \end{figure}
@@
 
-%\begin{figure}[h]
-%\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
-%\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data.
-The threshold in red assumes an $F_{ST}$ of 0.002.
-\label{fig:hist1}}
-%\end{figure}
-%
-%\begin{figure}[h]
-%	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
-%	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals.
-The threshold in red assumes an $F_{ST}$ of 0.002.
-\label{fig:hist2}}
-%\end{figure}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-country-UKBB}}
+	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
+\end{figure}
+
+\begin{figure}[htb]
+	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-country}}
+	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
+\end{figure}
 

### 2020-10-26 paper/paper-ancestry-matching.tex
last changes

@@
 \abstract{
-	Here, we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
+	Here we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
 	The proposed approach is particularly useful for methods that need to be applied in homogeneous samples.
@@
 The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric medians$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
-The (squared) distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
+The (squared) Euclidean distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
 In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1}, \ref{fig:compare-dist1}, \ref{fig:heatmap3} and \ref{fig:compare-dist3}).
 For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) and the four admixed American (AMR) populations are close to all European (EUR), South Asian (SAS) and African (AFR) populations when using the Bhattacharyya distance (Figure \ref{fig:heatmap1}).
-Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers.
-Note that we use 16 PCs for computing all PCA-based distances based on visual inspection of PC scores, as recommended in \cite{prive2020efficient}. 
+Using the shortest distances between pairs of individuals in two different populations is very sensitive to outliers.
+We also vary the number of PCs used for computing the Euclidean distances and how they compare with $F_{ST}$ in figure \ref{fig:compare-distK}.
 With 2 to 4 PCs, we are able to adequately separate distant populations, but not the closest ones.
 For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are superimposed (Figure \ref{fig:compare-distK}).
+When using more PCs (8, 16 or 25) to compute the distances, results remain similar.
 
@@
 We recall that this uses an automatic removal of LD when computing PCA and a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}.
-Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers (geometric medians$^\dag$) in the PCA space.
+Based on the results from the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers (geometric medians$^\dag$) in the PCA space.
 Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
@@
 Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
-Based on the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
+When using the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
 More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}).
@@
 Only 47 individuals were misclassified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
-However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).
+However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our genetic ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).
 This possible discrepancy between self-reported ancestry and genetic ancestry has been reported before \cite[]{mersha2015self}.
@@
 Finally, we show several ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
-The first approach is to simply match individuals that are close enough to one of the 1000G populations, as described previously.
-
+One first possible approach is to simply match individuals that are close enough to one of the 1000G populations, as described previously.
 Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth, e.g.\ available in the UK Biobank (Fields 21000 and 20115). 
@@
 In the UK Biobank data, we can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all centers).
-This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (TODO: TABLE).
+This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (Table \ref{tab:ancestry-groups}).
 Only 3.7\% of all individuals could not be matched.
-The resulting clusters are presented on the PCA space in figure \ref{fig:grouping}.
+The resulting clusters are presented in the PCA space in figure \ref{fig:grouping}.
 
@@
 Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals.
-Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
+Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset of centers with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
 Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
-Only 2.8\% of all individuals could not be matched (TODO TABLE).
-The resulting clusters are presented on the PCA space in figure \ref{fig:grouping2}.
+Only 2.8\% of all individuals could not be matched (Table \ref{tab:country-groups}).
+The resulting clusters are presented in the PCA space in figure \ref{fig:grouping2}.
 Note that these clusters probably include individuals from nearby countries as well.
@@
 This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
-Indeed, consider an example of an admixed individual of say 25\% European ancestry and 75\% African ancestry.
+Indeed, consider e.g.\ an admixed individual of say 25\% European ancestry and 75\% African ancestry.
 The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
@@
 Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve our proposed method.
-Nevertheless, we also show how to define homogeneous ancestry groups without using the 1000G data.
-When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, it is possible to directly restrict to a homogeneous subset of the data.
+Nevertheless, we also show how to define homogeneous ancestry groups without using the 1000G data, either by using self-reported ancestries or countries of birth.
+When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, we also show how to directly restrict to a homogeneous subset of the data.
 
@@
 Several studies have used only the first two PCs for ancestry inference.
-This is not enough to distinguish between Admixed Americans and South Asians (see AMR and SAS in figure \ref{fig:grouping}), let alone for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
+We have shown here that using two PCs (or even four) is not enough for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
 As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
-Moreover, we believe our proposed method to be robust to the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
+Moreover, we believe our proposed method to be robust to increasing the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
 As for the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-centers-1000G}}
-	\caption{Heatmap with clustering based on the distances between centers of pairs of the 26 1000G populations. \label{fig:heatmap2}}
+	\caption{Heatmap with clustering based on the Euclidean distances between centers of pairs of the 26 1000G populations. \label{fig:heatmap2}}
 \end{figure}
@@
 \centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
-\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations.
+\caption{Comparing $F_{ST}$ to the squared Euclidean distances on the PCA space between centers of pairs of the 26 1000G populations.
 Distances are computed using different numbers of Principal Components (PCs).
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-country-UKBB}}
+	\centerline{\includegraphics[width=0.9\textwidth]{heatmap-country-UKBB}}
 	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
 \end{figure}
+
+\FloatBarrier
+
+\input{../tables/country-groups}
 

### 2020-11-18 paper/paper-ancestry-matching-supp.tex
move to suppmat

--- 2020-10-26 paper/paper-ancestry-matching.tex
@@
 \begin{document}
-
-\title{Ancestry inference and grouping\\from principal component analysis of genetic data
-}
-\author{Florian Priv\'e$^{\text{1,}*}$}
-
-\date{~ }
-\maketitle
-
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
-
-\noindent Contact:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\end{itemize}
-
-\vspace*{5em}
-
-\abstract{
-	Here we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
-	The proposed approach is particularly useful for methods that need to be applied in homogeneous samples.
-	First, we show that Euclidean distances in the PCA space are proportional to $F_{ST}$ between populations.
-	Then, we show how to use this PCA-based distance to infer ancestry in the UK Biobank and the POPRES datasets.
-	We propose two solutions, either relying on projection of PCs to reference populations such as from the 1000 Genomes Project, or by directly using the internal data.
-	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-
-\subsection*{Introduction}
-
-Principal Component Analysis (PCA) has been widely used to correct for population structure in association studies and has been shown to mirror geography in Europe \cite[]{price2006principal,novembre2008genes}.
-Due to its popularity, many methods has been developed for efficiently performing PCA \cite[]{abraham2017flashpca2,prive2020efficient} as well as appropriately projecting samples onto a reference PCA space \cite[]{zhang2020fast,prive2020efficient}, making possible to perform these analyses for ever increasing datasets in human genetics.
-Naturally, PCA has also been used for ancestry inference. 
-However, among all studies where we have seen PCA used for ancestry inference, we have found there was no consensus on what is the most appropriate method for inferring ancestry using PCA.
-For example, there may be divergences on which distance metric to use and the number of PCs to use to compute these distances.
-
-Here, we first compare several distance metrics with the popular $F_{ST}$ statistic between populations.
-We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
-We do not provide a method to infer admixture coefficients nor local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
-However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of homogeneous ancestry.
-This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the population used for training \cite[]{martin2017human}.
-
-\vspace*{2em}
-$^\dag$ Further defined in supplementary section ``Defintions''.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{Measures of genetic dissimilarity between populations}
-
-We first compare four measures of genetic dissimilarity using populations of the 1000 Genomes Project (1000G$^\dag$, \cite{10002015global}). 
-The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
-We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}, and the clustering of these populations based on $F_{ST}$ in figure \ref{fig:heatmap0}.
-The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric medians$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
-The (squared) Euclidean distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
-In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1}, \ref{fig:compare-dist1}, \ref{fig:heatmap3} and \ref{fig:compare-dist3}).
-For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) and the four admixed American (AMR) populations are close to all European (EUR), South Asian (SAS) and African (AFR) populations when using the Bhattacharyya distance (Figure \ref{fig:heatmap1}).
-Using the shortest distances between pairs of individuals in two different populations is very sensitive to outliers.
-We also vary the number of PCs used for computing the Euclidean distances and how they compare with $F_{ST}$ in figure \ref{fig:compare-distK}.
-With 2 to 4 PCs, we are able to adequately separate distant populations, but not the closest ones.
-For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are superimposed (Figure \ref{fig:compare-distK}).
-When using more PCs (8, 16 or 25) to compute the distances, results remain similar.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{PCA-based ancestry inference}
-
-We project the dataset of interest onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
-We recall that this uses an automatic removal of LD when computing PCA and a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}.
-Based on the results from the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers (geometric medians$^\dag$) in the PCA space.
-Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
-This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
-When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest reference population center.
-
-We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
-These individuals seem to originate from many parts of the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
-Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
-When using the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
-More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}).
-This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries.
-Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within ``White and Black Africa'' and 93.0\% within ``White and Asian'' ancestries.
-Only 47 individuals were misclassified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}).
-However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our genetic ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).
-This possible discrepancy between self-reported ancestry and genetic ancestry has been reported before \cite[]{mersha2015self}.
-
-We also test the approach proposed in \cite{zhang2020fast} which consists in finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals.
-When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
-Less than 0.5\% could not be matched by their method (Table \ref{tab:ancestry-pred-kNN}).
-Of note, they could match much more admixed individuals, whereas they set a high probability threshold aiming at discarding such admixed individuals. 
-Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
-Finally, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-fine-pred}).
-We also applied our ancestry detection technique to the European individuals of the POPRES data \cite[]{nelson2008population}. 
-Only 16 out of the 1385 individuals (1.2\%) could not be matched, of which 11 were from East or South-East Europe (Table \ref{tab:ancestry-pred-popres}).
-Note that all individuals that we could match were identified as of European ancestry. 
-We could also identify accurately sub-regions of Europe; e.g.\ 261 out of 264 Spanish and Portugese individuals were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
-
-\begin{table}[htb]
-	\centering
-	\caption{Self-reported ancestry (left) of UKBB individuals and their matching to 1000G continental populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-	\label{tab:infer-UKBB-superpop}
-	\begin{tabular}{|l|c|c|c|c|c|c|}
-		\hline
-		& AFR & AMR & EAS & EUR & SAS & Not matched \\ 
-		\hline
-		British & 2 &  & 1 & 421457 & 6 & 9548 \\ 
-		Irish &  &  &  & 12328 &  & 425 \\ 
-		White & 1 & 1 & 1 & 499 &  & 43 \\ 
-		Other White &  & 40 &  & 11334 & 1 & 4440 \\ 
-		\hline
-		Indian &  &  &  & 5 & 4922 & 789 \\ 
-		Pakistani &  &  &  &  & 1421 & 327 \\ 
-		Bangladeshi &  &  &  &  & 217 & 4 \\ 
-		Chinese &  &  & 1453 & 1 &  & 50 \\ 
-		Other Asian & 1 &  & 279 &  & 939 & 528 \\ 
-		\hline
-		Caribbean & 3848 &  &  &  & 25 & 424 \\ 
-		African & 2633 &  &  & 1 &  & 570 \\ 
-		Other Black & 74 &  &  &  & 2 & 42 \\ 
-		\hline
-		Asian or Asian British &  &  & 2 &  & 20 & 20 \\ 
-		Black or Black British & 20 &  &  & 2 &  & 4 \\ 
-		White and Black Caribbean & 24 & 1 &  & 8 & 1 & 563 \\ 
-		White and Black African & 5 &  &  & 6 &  & 391 \\ 
-		White and Asian &  & 1 & 2 & 27 & 26 & 746 \\ 
-		\hline
-		Unknown  & 835 & 173 & 576 & 2296 & 633 & 3307 \\ 
-		\hline
-	\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{PCA-based ancestry grouping}
-
-Finally, we show several ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
-One first possible approach is to simply match individuals that are close enough to one of the 1000G populations, as described previously.
-Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth, e.g.\ available in the UK Biobank (Fields 21000 and 20115). 
-This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.
-In the UK Biobank data, we can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all centers).
-This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (Table \ref{tab:ancestry-groups}).
-Only 3.7\% of all individuals could not be matched.
-The resulting clusters are presented in the PCA space in figure \ref{fig:grouping}.
-
-One could do the same using the countries of birth instead of the self-reported ancestries.
-Again, the country of birth may sometime not reflect the ancestral origin.
-Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals.
-Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset of centers with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
-Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
-Only 2.8\% of all individuals could not be matched (Table \ref{tab:country-groups}).
-The resulting clusters are presented in the PCA space in figure \ref{fig:grouping2}.
-Note that these clusters probably include individuals from nearby countries as well.
-
-Finally, when we know that the data is composed of a predominant ancestry, we can define a single homogeneous cluster by simply restricting to individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
-When doing so, we can cluster 91\% of the data into one cluster composed of 421,871 British, 12,039 Irish, 8351 ``Other White'', 1814 individuals of unknown ancestry, 467 ``White'' and 41 individuals of other self-reported ancestries.
-This is made possible because we use the geometric median which is robust to outliers.
-
-\begin{figure}[htb]
-	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
-	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{Discussion}
-
-Here we propose a PCA-based method for ancestry inference and grouping individuals into genetically homogeneous clusters.
-We show how the PCA-based distance is related to the $F_{ST}$, which allows to compute distances based on PC scores directly.
-This relation between $F_{ST}$ and (squared) Euclidean distances in the PCA space has been previously shown for two populations only \cite[]{mcvean2009genealogical}.
-Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a single homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
-When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appears suboptimal here compared to using a simple Euclidean distance.
-We hypothesize that the main issue with this approach is that an admixed population covers a large volume in the PCA space, therefore all distances to this population cluster are small because of the covariance component from the Mahalanobis distance.
-In contrast, here we propose to directly use the global scale of the PC scores, which is invariant from the cluster scattering.
-This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
-Indeed, consider e.g.\ an admixed individual of say 25\% European ancestry and 75\% African ancestry.
-The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
-We also believe our proposed method to be more robust than machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
-In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
-
-Yet, our proposed method also has limitations. 
-First, since we match target individuals to 1000G populations, if individuals are far from all 26 1000G populations, then they would not be matched.
-When looking at the POPRES data, more individuals from East Europe could not be matched. 
-This is not surprising because there are no East European population in the 1000G data.
-Moreover, if we look at the location of the 1000G populations on a map, we can see that it lacks representation of many parts of the world (Figure \ref{fig:map}).
-This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
-Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve our proposed method.
-Nevertheless, we also show how to define homogeneous ancestry groups without using the 1000G data, either by using self-reported ancestries or countries of birth.
-When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, we also show how to directly restrict to a homogeneous subset of the data.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
-	\caption{
-		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
-Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
-		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
-		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
-		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
-\label{fig:map}}
-\end{figure}
-
-A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.
-Several studies have used only the first two PCs for ancestry inference.
-We have shown here that using two PCs (or even four) is not enough for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
-As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
-Moreover, we believe our proposed method to be robust to increasing the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
-As for the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
-Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
-This threshold can also be adjusted depending on how homogeneous one want each cluster to be.
-
-In conclusion, we believe our proposed approach to be a simple and robust way to infer global ancestry and to define groups of homogeneous ancestry.
-It is also very fast, allowing to infer ancestry for 488,371 individuals in 20 minutes using 16 cores.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\section*{Software and code availability}
-
-The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The code used in this paper is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.
-
-\section*{Acknowledgements}
-
-Authors thank Alex Diaz-Papkovich, Clive Hoggart and others for their useful feedback.
-This research has been conducted using the UK Biobank Resource under Application Number 41181.
-
-\section*{Funding}
-
-F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).
-
-\section*{Declaration of Interests}
-
-The authors declare no competing interests.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\newpage
-
-\bibliographystyle{natbib}
-\bibliography{refs}
 
@@
 
+\clearpage
+
+\bibliographystyle{natbib}
+\bibliography{refs}
+
 \end{document}

### 2020-11-18 paper/paper-ancestry-matching.tex
move to suppmat

@@
 
-%\usepackage{xr}
-%\externaldocument{SCT-supp}
+\usepackage{xr}
+\externaldocument{paper-ancestry-matching-supp}
 
@@
 \vspace*{2em}
-$^\dag$ Further defined in supplementary section ``Defintions''.
+$^\dag$ Further defined in supplementary section ``Definitions $\dag$ and methods''.
 
@@
 
-\clearpage
-
-\section*{Supplementary Materials}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-\renewcommand{\theequation}{S\arabic{equation}}
-\setcounter{equation}{0}
-\renewcommand{\thesection}{S\arabic{section}}
-\setcounter{section}{1}
-
-\subsection*{Definitions $\dag$ and methods}
-
-\begin{itemize}
-
-	\item The {\bf 1000 Genomes Project (1000G)} data is composed of approximately 100 individuals for each of 26 populations worldwide (described at \url{https://www.internationalgenome.org/category/population/}), including 7 African (AFR), 5 East Asian (EAS), 5 South Asian (SAS), 5 European (EUR) and 4 admixed American (AMR) populations.
-	Here we used the transformed data in PLINK format provided in \cite{prive2020efficient}.
-
-	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
-	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.
-
-	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix \cite[]{prive2020efficient}.
-They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.
-
-	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ is defined as
-	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
-	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}.
-	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.
-
-	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points.
-We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.
-
-	\item The {\bf UK Biobank} is a large cohort of half a million individuals from the UK, for which we have access to both genotypes and multiple phenotypes (\url{https://www.ukbiobank.ac.uk/}).
-	We apply some quality control filters to the genotyped data; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-	This results in 488,371 individuals and 504,139 genetic variants.
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\subsection*{Measures of genetic dissimilarity between populations}
-
-\input{../tables/fst-values-AFR}
-\input{../tables/fst-values-AMR}
-\input{../tables/fst-values-EAS}
-\input{../tables/fst-values-EUR}
-\input{../tables/fst-values-SAS}
-
-\FloatBarrier
-
-\begin{figure}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{heatmap-Fst-1000G}}
-\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}. \label{fig:heatmap0}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-bhattacharyya-1000G}}
-	\caption{Heatmap with clustering based on the Bhattacharyya distances between pairs of the 26 1000G populations. \label{fig:heatmap1}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\textwidth]{compare-Bhattacharyya-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the Bhattacharyya distance on the PCA space between pairs of the 26 1000G populations. \label{fig:compare-dist1}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-centers-1000G}}
-	\caption{Heatmap with clustering based on the Euclidean distances between centers of pairs of the 26 1000G populations. \label{fig:heatmap2}}
-\end{figure}
-
-\begin{figure}[h]
-\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
-\caption{Comparing $F_{ST}$ to the squared Euclidean distances on the PCA space between centers of pairs of the 26 1000G populations.
-Distances are computed using different numbers of Principal Components (PCs).
-\label{fig:compare-distK}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-closest-1000G}}
-	\caption{Heatmap with clustering based on the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:heatmap3}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\textwidth]{compare-closest-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:compare-dist3}}
-\end{figure}
-
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{PCA-based ancestry inference}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{proj-UKBB-to-1000G}}
-	\caption{First 18 PC scores of the 1000G data (in black), onto which the UK Biobank data has been projected (in red). \label{fig:proj-UKBB}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-mismatch}}
-	\caption{PC scores (computed in the UK Biobank) colored by self-reported ancestry.
-On the left, these are 50,000 random individuals.
-On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and our ancestry estimation.
-\label{fig:mismatch}}
-\end{figure}
-
-\input{../tables/matched-pop-KNN}
-\input{../tables/infer-UKBB-pop}
-\input{../tables/POPRES-matching}
-
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{PCA-based ancestry grouping}
-
-\input{../tables/ancestry-groups}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.9\textwidth]{heatmap-country-UKBB}}
-	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
-\end{figure}
-
-\FloatBarrier
-
-\input{../tables/country-groups}
-
-\begin{figure}[htb]
-	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-country}}
-	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
-	\caption{Histogram of (log) squared distances from the UK Biobank PC scores to the geometric median of the all UKBB individuals.
-Here we use a threshold at 7, based on visual inspection.
-Alternatively, a more stringent threshold at 6 could also be used.
-\label{fig:hist3}}
-\end{figure}
-
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2021-01-12 paper/supp-note-ancestry-grouping.tex
make as supp note

--- 2020-10-26 paper/paper-ancestry-matching.tex
@@
 
-%\usepackage{xr}
-%\externaldocument{SCT-supp}
-
 %\linenumbers
@@
 %\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage{natbib}
+\bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 
-\title{Ancestry inference and grouping\\from principal component analysis of genetic data
+\title{Supplementary Note: Ancestry inference and grouping from principal component analysis of genetic data
 }
-\author{Florian Priv\'e$^{\text{1,}*}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
-
-\noindent Contact:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\end{itemize}
-
-\vspace*{5em}
-
-\abstract{
-	Here we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
-	The proposed approach is particularly useful for methods that need to be applied in homogeneous samples.
-	First, we show that Euclidean distances in the PCA space are proportional to $F_{ST}$ between populations.
-	Then, we show how to use this PCA-based distance to infer ancestry in the UK Biobank and the POPRES datasets.
-	We propose two solutions, either relying on projection of PCs to reference populations such as from the 1000 Genomes Project, or by directly using the internal data.
-	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
-}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-
-\subsection*{Introduction}
 
@@
 \vspace*{2em}
-$^\dag$ Further defined in supplementary section ``Defintions''.
+$^\dag$ Further defined in section ``Definitions $\dag$ and methods''.
 
@@
 
-\begin{table}[htb]
-	\centering
-	\caption{Self-reported ancestry (left) of UKBB individuals and their matching to 1000G continental populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-	\label{tab:infer-UKBB-superpop}
-	\begin{tabular}{|l|c|c|c|c|c|c|}
-		\hline
-		& AFR & AMR & EAS & EUR & SAS & Not matched \\ 
-		\hline
-		British & 2 &  & 1 & 421457 & 6 & 9548 \\ 
-		Irish &  &  &  & 12328 &  & 425 \\ 
-		White & 1 & 1 & 1 & 499 &  & 43 \\ 
-		Other White &  & 40 &  & 11334 & 1 & 4440 \\ 
-		\hline
-		Indian &  &  &  & 5 & 4922 & 789 \\ 
-		Pakistani &  &  &  &  & 1421 & 327 \\ 
-		Bangladeshi &  &  &  &  & 217 & 4 \\ 
-		Chinese &  &  & 1453 & 1 &  & 50 \\ 
-		Other Asian & 1 &  & 279 &  & 939 & 528 \\ 
-		\hline
-		Caribbean & 3848 &  &  &  & 25 & 424 \\ 
-		African & 2633 &  &  & 1 &  & 570 \\ 
-		Other Black & 74 &  &  &  & 2 & 42 \\ 
-		\hline
-		Asian or Asian British &  &  & 2 &  & 20 & 20 \\ 
-		Black or Black British & 20 &  &  & 2 &  & 4 \\ 
-		White and Black Caribbean & 24 & 1 &  & 8 & 1 & 563 \\ 
-		White and Black African & 5 &  &  & 6 &  & 391 \\ 
-		White and Asian &  & 1 & 2 & 27 & 26 & 746 \\ 
-		\hline
-		Unknown  & 835 & 173 & 576 & 2296 & 633 & 3307 \\ 
-		\hline
-	\end{tabular}
-\end{table}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-One could do the same using the countries of birth instead of the self-reported ancestries.
+{\bf One could do the same using the countries of birth instead of the self-reported ancestries, which we use in the main text.}
 Again, the country of birth may sometime not reflect the ancestral origin.
@@
 
+\input{../tables/country-groups}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{UKBB-matched-country}}
+	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
+\end{figure}
+
 Finally, when we know that the data is composed of a predominant ancestry, we can define a single homogeneous cluster by simply restricting to individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
@@
 This is made possible because we use the geometric median which is robust to outliers.
-
-\begin{figure}[htb]
-	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
-	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
-\end{figure}
 
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
-	\caption{
-		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
-Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
-		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
-		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
-		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
-\label{fig:map}}
-\end{figure}
-
 A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\section*{Software and code availability}
-
-The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-The code used in this paper is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.
-
-\section*{Acknowledgements}
-
-Authors thank Alex Diaz-Papkovich, Clive Hoggart and others for their useful feedback.
-This research has been conducted using the UK Biobank Resource under Application Number 41181.
-
-\section*{Funding}
-
-F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).
-
-\section*{Declaration of Interests}
-
-The authors declare no competing interests.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\newpage
-
-\bibliographystyle{natbib}
-\bibliography{refs}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
+	\caption{
+		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
+Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
+		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
+		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
+		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
+\label{fig:map}}
+\end{figure}
 
@@
 
-\section*{Supplementary Materials}
+\subsection*{Definitions $\dag$ and methods}
 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-\renewcommand{\theequation}{S\arabic{equation}}
-\setcounter{equation}{0}
-\renewcommand{\thesection}{S\arabic{section}}
-\setcounter{section}{1}
-
-\subsection*{Definitions $\dag$ and methods}
+Note that the code used in this supplementary note is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.
 
@@
 
+\bibliographystyle{natbib}
+\bibliography{refs}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
+\section*{Additional Figures and Tables}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
+\renewcommand{\theequation}{S\arabic{equation}}
+\setcounter{equation}{0}
+\renewcommand{\thesection}{S\arabic{section}}
+\setcounter{section}{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection*{Measures of genetic dissimilarity between populations}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-Fst-1000G}}
-\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}. \label{fig:heatmap0}}
+	\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations.
+Corresponding values are reported in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}.
+\label{fig:heatmap0}}
 \end{figure}
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{proj-UKBB-to-1000G}}
+	\centerline{\includegraphics[width=0.9\textwidth]{proj-UKBB-to-1000G}}
 	\caption{First 18 PC scores of the 1000G data (in black), onto which the UK Biobank data has been projected (in red). \label{fig:proj-UKBB}}
 \end{figure}
+
+\FloatBarrier
+
+\begin{table}[htb]
+	\centering
+	\caption{Self-reported ancestry (left) of UKBB individuals and their matching to 1000G continental populations (top) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+	\label{tab:infer-UKBB-superpop}
+	\begin{tabular}{|l|c|c|c|c|c|c|}
+		\hline
+		& AFR & AMR & EAS & EUR & SAS & Not matched \\ 
+		\hline
+		British & 2 &  & 1 & 421457 & 6 & 9548 \\ 
+		Irish &  &  &  & 12328 &  & 425 \\ 
+		White & 1 & 1 & 1 & 499 &  & 43 \\ 
+		Other White &  & 40 &  & 11334 & 1 & 4440 \\ 
+		\hline
+		Indian &  &  &  & 5 & 4922 & 789 \\ 
+		Pakistani &  &  &  &  & 1421 & 327 \\ 
+		Bangladeshi &  &  &  &  & 217 & 4 \\ 
+		Chinese &  &  & 1453 & 1 &  & 50 \\ 
+		Other Asian & 1 &  & 279 &  & 939 & 528 \\ 
+		\hline
+		Caribbean & 3848 &  &  &  & 25 & 424 \\ 
+		African & 2633 &  &  & 1 &  & 570 \\ 
+		Other Black & 74 &  &  &  & 2 & 42 \\ 
+		\hline
+		Asian or Asian British &  &  & 2 &  & 20 & 20 \\ 
+		Black or Black British & 20 &  &  & 2 &  & 4 \\ 
+		White and Black Caribbean & 24 & 1 &  & 8 & 1 & 563 \\ 
+		White and Black African & 5 &  &  & 6 &  & 391 \\ 
+		White and Asian &  & 1 & 2 & 27 & 26 & 746 \\ 
+		\hline
+		Unknown  & 835 & 173 & 576 & 2296 & 633 & 3307 \\ 
+		\hline
+	\end{tabular}
+\end{table}
 
@@
 On the left, these are 50,000 random individuals.
-On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and our ancestry estimation.
+On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and our ancestry estimation (see table \ref{tab:infer-UKBB-superpop}).
 \label{fig:mismatch}}
@@
 
-\input{../tables/ancestry-groups}
-
 \begin{figure}[h]
@@
 
-\FloatBarrier
-
-\input{../tables/country-groups}
+\input{../tables/ancestry-groups}
 
 \begin{figure}[htb]
-	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-country}}
-	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
+	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-matched-ancestry}}
+	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
 \end{figure}
+
+\FloatBarrier
 
@@
 
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2021-01-14 paper/supp-note-ancestry-grouping.tex
restructure supp note

@@
 
-\title{Supplementary Note: Ancestry inference and grouping from principal component analysis of genetic data
-}
+\title{Supplementary Note: Ancestry inference and grouping from principal component analysis of genetic data}
 
@@
 
-Principal Component Analysis (PCA) has been widely used to correct for population structure in association studies and has been shown to mirror geography in Europe \cite[]{price2006principal,novembre2008genes}.
-Due to its popularity, many methods has been developed for efficiently performing PCA \cite[]{abraham2017flashpca2,prive2020efficient} as well as appropriately projecting samples onto a reference PCA space \cite[]{zhang2020fast,prive2020efficient}, making possible to perform these analyses for ever increasing datasets in human genetics.
-Naturally, PCA has also been used for ancestry inference. 
-However, among all studies where we have seen PCA used for ancestry inference, we have found there was no consensus on what is the most appropriate method for inferring ancestry using PCA.
-For example, there may be divergences on which distance metric to use and the number of PCs to use to compute these distances.
-
-Here, we first compare several distance metrics with the popular $F_{ST}$ statistic between populations.
-We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
-We do not provide a method to infer admixture coefficients nor local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
-However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of homogeneous ancestry.
-This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the population used for training \cite[]{martin2017human}.
-
-\vspace*{2em}
+\vspace*{1em}
 $^\dag$ Further defined in section ``Definitions $\dag$ and methods''.
+\vspace*{1em}
 
@@
 The (squared) Euclidean distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
-In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1}, \ref{fig:compare-dist1}, \ref{fig:heatmap3} and \ref{fig:compare-dist3}).
+This relation between $F_{ST}$ and (squared) Euclidean distances in the PCA space has been previously shown for two populations only \cite[]{mcvean2009genealogical}.
+
+\begin{figure}[htb]
+	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
+	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
+\end{figure}
+
+Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a single homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
+When looking at distances between two populations, this corresponds to using the Bhattacharyya distance.
+However, in contrast to Euclidean distances, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1}, \ref{fig:compare-dist1}, \ref{fig:heatmap3} and \ref{fig:compare-dist3}).
 For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) and the four admixed American (AMR) populations are close to all European (EUR), South Asian (SAS) and African (AFR) populations when using the Bhattacharyya distance (Figure \ref{fig:heatmap1}).
-Using the shortest distances between pairs of individuals in two different populations is very sensitive to outliers.
+We hypothesize that the main issue with this approach is that an admixed population covers a large volume in the PCA space, therefore all distances to this population cluster are small because of the covariance component from the Mahalanobis distance.
+In contrast, the global scale of the PC scores used when using Euclidean distances is invariant from the cluster scattering.
+
 We also vary the number of PCs used for computing the Euclidean distances and how they compare with $F_{ST}$ in figure \ref{fig:compare-distK}.
@@
 When using more PCs (8, 16 or 25) to compute the distances, results remain similar.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
-\end{figure}
 
@@
 We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
+For 488,371 individuals, this procedure takes less than 20 minutes using 16 cores.
 These individuals seem to originate from many parts of the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
@@
 Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
+The global scale used in Euclidean distances makes it more robust to infer ancestry as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
+Indeed, consider e.g.\ an admixed individual of say 25\% European ancestry and 75\% African ancestry.
+The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
+We also believe our proposed method to be more robust than machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
+In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
+
 Finally, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-fine-pred}).
@@
 We could also identify accurately sub-regions of Europe; e.g.\ 261 out of 264 Spanish and Portugese individuals were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
+
+The proposed method has two possible limitations. 
+First, since we match target individuals to 1000G populations, if individuals are far from all 26 1000G populations, then they would not be matched.
+When looking at the POPRES data, more individuals from East Europe could not be matched. 
+This is not surprising because there are no East European population in the 1000G data.
+Moreover, if we look at the location of the 1000G populations on a map, we can see that it lacks representation of many parts of the world (Figure \ref{fig:map}).
+This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
+Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve the proposed method.
+A second potential limitation of the proposed method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.
+Several studies have used only the first two PCs for ancestry inference.
+We have shown here that using two PCs (or even four) is not enough for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
+As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
+Moreover, we believe our proposed method to be robust to increasing the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
+As for the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
+Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
+This threshold can also be adjusted depending on how homogeneous one want each cluster to be.
 
@@
 Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth, e.g.\ available in the UK Biobank (Fields 21000 and 20115). 
-This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.
+{\bf This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.}
 In the UK Biobank data, we can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all centers).
@@
 Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset of centers with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
-Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
-Only 2.8\% of all individuals could not be matched (Table \ref{tab:country-groups}).
+Based on the previous clustering and the available sample sizes, we choose to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
+Only 2.8\% of all individuals could not be matched to one of these eight groups (Table \ref{tab:country-groups}).
 The resulting clusters are presented in the PCA space in figure \ref{fig:grouping2}.
@@
 
-\subsection*{Discussion}
-
-Here we propose a PCA-based method for ancestry inference and grouping individuals into genetically homogeneous clusters.
-We show how the PCA-based distance is related to the $F_{ST}$, which allows to compute distances based on PC scores directly.
-This relation between $F_{ST}$ and (squared) Euclidean distances in the PCA space has been previously shown for two populations only \cite[]{mcvean2009genealogical}.
-Previously, we and others proposed to use (robust) Mahalanobis distances to infer ancestry or identify a single homogeneous group of individuals \cite[]{peterson2017utility,prive2020efficient}. 
-When looking at distances between two populations, this corresponds to using the Bhattacharyya distance, which appears suboptimal here compared to using a simple Euclidean distance.
-We hypothesize that the main issue with this approach is that an admixed population covers a large volume in the PCA space, therefore all distances to this population cluster are small because of the covariance component from the Mahalanobis distance.
-In contrast, here we propose to directly use the global scale of the PC scores, which is invariant from the cluster scattering.
-This global scale makes it also more robust to infer ancestry with our method as compared to using relative proportions from k=20 nearest neighbors (kNN, \cite{zhang2020fast}).
-Indeed, consider e.g.\ an admixed individual of say 25\% European ancestry and 75\% African ancestry.
-The kNN-based method is likely to identify this individual as of African ancestry, while our method will probably be unable to match it, which is a beneficial feature when we are interested in defining genetically homogeneous groups.
-We also believe our proposed method to be more robust than machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
-In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
-
-Yet, our proposed method also has limitations. 
-First, since we match target individuals to 1000G populations, if individuals are far from all 26 1000G populations, then they would not be matched.
-When looking at the POPRES data, more individuals from East Europe could not be matched. 
-This is not surprising because there are no East European population in the 1000G data.
-Moreover, if we look at the location of the 1000G populations on a map, we can see that it lacks representation of many parts of the world (Figure \ref{fig:map}).
-This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
-Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve our proposed method.
-Nevertheless, we also show how to define homogeneous ancestry groups without using the 1000G data, either by using self-reported ancestries or countries of birth.
-When a predominant genetic ancestry is present in the data, such as British in the UK Biobank \cite[]{bycroft2017genome} or Danish in the iPSYCH data \cite[]{pedersen2018ipsych2012}, we also show how to directly restrict to a homogeneous subset of the data.
-
-A second potential limitation of our method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.
-Several studies have used only the first two PCs for ancestry inference.
-We have shown here that using two PCs (or even four) is not enough for distinguishing between populations at the sub-continental level (Figure \ref{fig:compare-distK}).
-As in \cite{prive2020efficient}, we recommend to use all PCs that visually separate some populations.
-Moreover, we believe our proposed method to be robust to increasing the number of PCs used because contribution to the Euclidean distance is smaller for later PCs than for first PCs.
-As for the distance limit, we have shown here how to define it to approximately correspond to an $F_{ST}$ of 0.002.
-Alternatively, a threshold can be chosen based on the visual inspection of the histogram of distances (on a log scale).
-This threshold can also be adjusted depending on how homogeneous one want each cluster to be.
-
-In conclusion, we believe our proposed approach to be a simple and robust way to infer global ancestry and to define groups of homogeneous ancestry.
-It is also very fast, allowing to infer ancestry for 488,371 individuals in 20 minutes using 16 cores.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
-	\caption{
-		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
-Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
-		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
-		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
-		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
-\label{fig:map}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
+%\clearpage
 
@@
 
-\clearpage
+%\clearpage
 
@@
 
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
+	\caption{
+		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115).
+Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
+		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
+		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
+		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location.
+\label{fig:map}}
+\end{figure}
+
 \FloatBarrier
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.9\textwidth]{heatmap-country-UKBB}}
-	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
-\end{figure}
-
 \input{../tables/ancestry-groups}
@@
 	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping}}
+\end{figure}
+
+\FloatBarrier
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.9\textwidth]{heatmap-country-UKBB}}
+	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
 \end{figure}

### 2021-02-19 paper/supp-note-ancestry-grouping.tex
save

@@
 We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}, and the clustering of these populations based on $F_{ST}$ in figure \ref{fig:heatmap0}.
-The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric medians$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
+The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric medians$^\dag$) of the two populations; and 3) the shortest distance between pairs of PC scores, one from each of the two populations.
 The (squared) Euclidean distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
@@
 For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are superimposed (Figure \ref{fig:compare-distK}).
-When using more PCs (8, 16 or 25) to compute the distances, results remain similar.
+When using more PCs (8, 16 or 25) to compute the distances, results remain mostly similar.
 
@@
 We also believe our proposed method to be more robust than machine learning methods, because a machine learning method would try e.g.\ to differentiate between GBR and CEU 1000G populations, which are two very close populations of Northwest Europe ($F_{ST}$ of 0.0002).
-In other words, our distance-based method should benefit from the inclusion of any new reference population while it would make it increasingly complex to apply machine learning methods.
+In other words, our distance-based method should benefit from the inclusion of any new reference population, whereas it would make it increasingly complex to apply machine learning methods.
 
@@
 This issue has also been reported e.g.\ for Asian populations \cite[]{lu2013principal}.
-Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, which would also improve the proposed method.
+Therefore more diverse populations should be aggregated to better cover the worldwide genome diversity, such as with the Simons Genome Diversity Project \cite[]{mallick2016simons}, which would also improve the proposed method.
 A second potential limitation of the proposed method is that it has two hyper-parameters: the number of PCs used to compute the distances and the threshold on the minimum distance to any cluster center above which the ancestry is not matched.

### 2021-05-17 paper/supp-note-ancestry-grouping.tex
update

@@
 The (squared) Euclidean distance between population centers appears to be an appropriate PCA-based distance as it is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist2}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
+However, future work is needed to understand why residuals are bimodal for large distances (e.g.\ in figure \ref{fig:compare-dist2}).
 This relation between $F_{ST}$ and (squared) Euclidean distances in the PCA space has been previously shown for two populations only \cite[]{mcvean2009genealogical}.
@@
 	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
-	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
+	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space (i.e.\ using PC scores$^\dag$) between centers of pairs of the 26 1000G populations. \label{fig:compare-dist2}}
 \end{figure}
@@
 We recall that this uses an automatic removal of LD when computing PCA and a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}.
-Based on the results from the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers (geometric medians$^\dag$) in the PCA space.
+Based on the results from the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the Euclidean distance to these reference population centers in the PCA space (geometric medians$^\dag$ of PC scores$^\dag$).
 Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002.
@@
 Note that these clusters probably include individuals from nearby countries as well.
+Moreover, more clusters could probably be defined, e.g.\ the individuals with large values for PC6 in figure \ref{fig:grouping2} seem to originate from South America with many people from Colombia, Chile, Mexico, Peru, Ecuador, Venezuela, Bolivia, Brazil, and Argentina.
+However, here we decide to restrict to large enough clusters (e.g.\ with more than 1000 individuals).
+The cluster with small values for PC4 corresponds to Ashkenazi ancestry, and is described in the main text.
 
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{UKBB-matched-country}}
-	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
+	\caption{The first eight PC scores$^\dag$ computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
 \end{figure}
